---
title: "IVMC analysis and plots"
date:  "`r Sys.Date()`"
date-format: iso
format: 
    html:
        embed-resources: true
        toc: true
        toc-depth: 6
---
 
# Description

This script will generate the following plots:

- Fig 1 D, E, F, G
- Fig 3 C, G
- Fig 5 C

- Supplementary Fig 1 C, D, E, F
- Supplementary Fig 2 A
- Supplementary Fig 3 A, C, F, G, I
- Supplementary Fig 5 F


# preparation

## load libraries

```{r message=FALSE}
#| label: libraries
library(sva)
library(edgeR) 
library(dplyr)
library(PCAtools)
library(ComplexHeatmap)
library(gprofiler2)
gprofiler2::set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e112_eg59_p19")
```

## new folders
```{r}
#| label: folders
plot_dir <- "paper_figures_IVMC"
if (!file.exists(plot_dir)) {
    dir.create(plot_dir, showWarnings = FALSE, recursive = FALSE)
}

table_dir <- "paper_tables_IVMC"
if (!file.exists(table_dir)) {
    dir.create(table_dir, showWarnings = FALSE, recursive = FALSE)
}
```

## reading raw counts and tpm values
```{r}
#| label: data
all_counts <- read.delim("data/IVMC.raw_counts.20250403.tab")
# calculate cpm values on the complete data set first
all_counts.cpm <- cpm(all_counts, log=FALSE)

all_tpm <- read.delim("data/IVMC.TPM.20250403.tab")
```

## set up metadata 
```{r}
#| label: metadata
colnames(all_counts)
metadata <- data.frame(sample = colnames(all_counts),
                       donor = unlist(strsplit(colnames(all_counts),"_"))[rep(c(TRUE,FALSE,FALSE,FALSE),114)],
                       timepoint = unlist(strsplit(colnames(all_counts),"_"))[rep(c(FALSE,TRUE,FALSE,FALSE),114)],
                       breakdown = unlist(strsplit(colnames(all_counts),"_"))[rep(c(FALSE,FALSE,TRUE,FALSE),114)],
                       hormones  = unlist(strsplit(colnames(all_counts),"_"))[rep(c(FALSE,FALSE,FALSE,TRUE),114)] )
rownames(metadata) <- metadata$sample
metadata$timepoint_B_H <- paste(metadata$timepoint,metadata$breakdown,metadata$hormones, sep = "_")
metadata
```

## color and description scheme
```{r}
#| label: colors
source("../params.R")


timepoint_colors <- bulk_timepoint_colors[c("tp0","tp1","tp2","tp3","tp4","tp5","tp6")]
timepoint_colors.alt_nlb <- bulk_timepoint_colors[c("Pre-diff","P4-diff","H. W. 24h","H. W. 48h","P. B. 24h","E2-diff 24h","E2-diff 48h")]

heatmap_labels_colors <- c("black","white","black","black","white","black","black")

protocol_colors <- hormones_colors
names(protocol_colors) <- c("B_NH","IVMC")
```

```{r}
#| label: descriptions for timepoints
metadata$description <- bulk_timepoint_map_sl["tp0"]
metadata$description[metadata$timepoint == "tp1"] <- bulk_timepoint_map_sl["tp1"]
metadata$description[metadata$timepoint == "tp2"] <- bulk_timepoint_map_sl["tp2"]
metadata$description[metadata$timepoint == "tp3"] <- bulk_timepoint_map_sl["tp3"]
metadata$description[metadata$timepoint == "tp4"] <- bulk_timepoint_map_sl["tp4"]
metadata$description[metadata$timepoint == "tp5"] <- bulk_timepoint_map_sl["tp5"]
metadata$description[metadata$timepoint == "tp6"] <- bulk_timepoint_map_sl["tp6"]
# alternative2
metadata$description.alt2 <- bulk_timepoint_map_mlm["tp0"]
metadata$description.alt2[metadata$timepoint == "tp1"] <- bulk_timepoint_map_mlm["tp1"]
metadata$description.alt2[metadata$timepoint == "tp2"] <- bulk_timepoint_map_mlm["tp2"]
metadata$description.alt2[metadata$timepoint == "tp3"] <- bulk_timepoint_map_mlm["tp3"]
metadata$description.alt2[metadata$timepoint == "tp4"] <- bulk_timepoint_map_mlm["tp4"]
metadata$description.alt2[metadata$timepoint == "tp5"] <- bulk_timepoint_map_mlm["tp5"]
metadata$description.alt2[metadata$timepoint == "tp6"] <- bulk_timepoint_map_mlm["tp6"]
# alternative no line break
metadata$description.alt_nlb <- gsub("\n", " ", bulk_timepoint_map_mls["tp0"])
metadata$description.alt_nlb[metadata$timepoint == "tp1"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp1"])
metadata$description.alt_nlb[metadata$timepoint == "tp2"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp2"])
metadata$description.alt_nlb[metadata$timepoint == "tp3"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp3"])
metadata$description.alt_nlb[metadata$timepoint == "tp4"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp4"])
metadata$description.alt_nlb[metadata$timepoint == "tp5"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp5"])
metadata$description.alt_nlb[metadata$timepoint == "tp6"] <- gsub("\n", " ", bulk_timepoint_map_mls["tp6"])
# alternative two line breaks
metadata$description.alt_2lb[metadata$timepoint == "tp0"] <- "Pre-diff"
metadata$description.alt_2lb[metadata$timepoint == "tp1"] <- "P4-diff"
metadata$description.alt_2lb[metadata$timepoint == "tp2"] <- "Horm\nWithdr\n24h"
metadata$description.alt_2lb[metadata$timepoint == "tp3"] <- "Horm\nWithdr\n48h"
metadata$description.alt_2lb[metadata$timepoint == "tp4"] <- "Post-\nbreakdown\n24h"
metadata$description.alt_2lb[metadata$timepoint == "tp5"] <- "E2-diff\n24h"
metadata$description.alt_2lb[metadata$timepoint == "tp6"] <- "E2-diff\n48h"


# Pre-diff
metadata$timeline <- "other"
metadata$timeline[metadata$timepoint == "tp0"] <- "Pre-diff"
```

## batch correction by donor using ComBat-seq
```{r}
#| label: batch correction
# 'ComBat-seq()' requires integers 
forceMatrixToInteger <- function(m){
    apply (m, c (1, 2), function (x) {
         (as.integer(x))
    })
}

all_counts.integer <- forceMatrixToInteger(as.matrix(all_counts))
row.names(all_counts.integer) <- row.names(all_counts)

# actual batch correction
all_counts.CBSeq_byD <- ComBat_seq(all_counts.integer, batch=metadata$donor, group=NULL)
row.names(all_counts.CBSeq_byD) <- row.names(all_counts)

# calculate cpm values on the complete data set first
all_counts.bc.cpm <- cpm(all_counts.CBSeq_byD, log=FALSE)
```

## set up DGE
only one design matrix for Figures: 1F, S3A, S3C, S3F, S3I
```{r}
#| label: DGE
groups_all <- metadata$timepoint_B_H
donor <- metadata$donor
# generate DGEList object (we work with the raw counts)
counts.all.asDGEList <- DGEList(counts=all_counts,
                                     genes=rownames(all_counts), 
                                     group = groups_all )

counts.all.asDGEList

# filter lowly expressed genes 
# we have 6 replicas (aka donors)
mincpm <- 1
minsamples <- 6

counts.all.asDGEList <- counts.all.asDGEList[ rowSums(all_counts.cpm > mincpm) >= minsamples, , keep.lib.size=FALSE] 
dim(counts.all.asDGEList)

# normalization
counts.all.asDGEList <- calcNormFactors(counts.all.asDGEList)   
head(counts.all.asDGEList$samples)

plotMDS(counts.all.asDGEList)
```

## set up design matrix
(B032_tp0_NB_NH is used as intercept)
```{r}
#| label: design matrix
# use model.matrix
design_all <- model.matrix(~donor +group, data = counts.all.asDGEList$samples)
colnames(design_all)[2:24] <- substring(colnames(design_all)[2:24],6,15)
design_all
```

## dispersion estimation and fitting
```{r}
#| label: dispersion estimation and fitting
# dispersion estimation
counts.all.asDGEList <- estimateDisp(counts.all.asDGEList, design_all)
counts.all.asDGEList$common.dispersion

plotBCV(counts.all.asDGEList)

# fitting
fit_all <- glmQLFit(counts.all.asDGEList, design_all)

plotQLDisp(fit_all)
```


# Fig 1 D
PCA of batch corrected samples in the IVMC protocol analysed by bulk RNA sequencing, coloured by timepoint (n=6 independent organoid lines). 

## using B0xx_tp0_NB_NH, B0xx_tp1_NB_H, B0xx_tp2_NB_H, B0xx_tp3_NB_H, B0xx_tp4_B_H, B0xx_tp5_B_H, B0xx_tp6_B_H => "sel_IVMC"
```{r}
#| label: Fig 1 D
metadata.sel_IVMC <- mutate(as.data.frame(metadata), sel_IVMC = case_when(
                                timepoint == "tp0" ~ TRUE,
                                timepoint == "tp1" & hormones == "H" ~ TRUE,
                                timepoint == "tp2" & hormones == "H" ~ TRUE,
                                timepoint == "tp3" & hormones == "H" ~ TRUE,
                                timepoint == "tp4" & hormones == "H" & breakdown == "B" ~ TRUE,
                                timepoint == "tp5" & hormones == "H" & breakdown == "B" ~ TRUE,
                                timepoint == "tp6" & hormones == "H" & breakdown == "B" ~ TRUE,
                                TRUE ~ FALSE))

sel_IVMC <- metadata$sample[metadata.sel_IVMC$sel_IVMC]
sel_IVMC

counts.sel_IVMC<- all_counts[,sel_IVMC]
metadata.sel_IVMC <- metadata[metadata.sel_IVMC$sel_IVMC,]

metadata.sel_IVMC$description.alt_nlb <- factor(metadata.sel_IVMC$description.alt_nlb,
                                                levels = unique(metadata.sel_IVMC$description.alt_nlb))

all_counts.CBSeq_byD.sel_IVMC <- all_counts.CBSeq_byD[,sel_IVMC]

all_counts.CBSeq_byD.sel_IVMC.log.cpm.pca <- pca(cpm(all_counts.CBSeq_byD.sel_IVMC, log=TRUE), 
                              metadata = metadata.sel_IVMC, removeVar = 0.1)

xlab = paste("PC1, ",round(all_counts.CBSeq_byD.sel_IVMC.log.cpm.pca$variance["PC1"],digits = 2),"% variance",sep = "")
ylab = paste("PC2, ",round(all_counts.CBSeq_byD.sel_IVMC.log.cpm.pca$variance["PC2"],digits = 2),"% variance",sep = "")

plot_object<- biplot(all_counts.CBSeq_byD.sel_IVMC.log.cpm.pca, 
       colby = 'description.alt_nlb',
       colkey = timepoint_colors.alt_nlb,
       pointSize = 8,
       lab = NULL,
       legendPosition = 'right',
       legendIconSize = 8,
       colLegendTitle = " ",
       gridlines.major = FALSE,
       gridlines.minor = FALSE,
       xlab = xlab,
       ylab = ylab,
       title = "")

plot_object

file_name <- paste(plot_dir,"/Figure_1_D.pdf", sep = "")
pdf(file_name, width = 8, height = 5)
    plot_object
dev.off()
```


# Fig 1 E
Heatmap depicting batch corrected centered and scaled counts per million (cpm) of selected genes across the IVMC protocol
```{r}
genes <- c(#proliferation 
           "MKI67", "PCNA", "MCM2", "MCM7", "E2F2", 
           #stem cells 
           "SOX9", "FUT4","AXIN2", "CDH2", "ALDH1A1", "LGR5", 
           #inflammation 
           "IL1A","IL1B","PTGS1","PTGS2", 
           #diff 
           "SULT1E1", "GATA2","ALPP","KRT17",'IL6','ENPP3','LIF', "HEY1","NKD2", 
           #secretory 
           "SCGB1D2", "SCGB1D4", "SCGB2A1","SCGB2A2", "S100P", "PAEP", "SPP1", "TFF1", 
           #ciliated 
           "FOXJ1", "PIFO","RSPH1", "TPPP3", 
           #hypoxia 
           "EGLN3", "HIF3A","SLC2A1","HK2","VEGFA","ANGPTL4","HIF1A",  
           #angiogenesis 
           "ANG", "FGF2", "NOS2", "CXCL8" , "ADAMTS9" , 
           #fibrinolysis 
           "THBS1", "PLAUR", "PLAU", "PLAT",  "HAS3", 
           #immune recruitement 
           "CSF2", "CXCL1", "CXCL2", "CXCL5", "CXCL6",'IL23A', 
           #resolution of inflammation
           "ANXA1","IL1RN", "IL10","IL19", "DMBT1", "BCL2A1", 
           #wound healing 
           "ITGA2", "KRT5","KRT6A","KRT13", "EREG", "AREG", "HBEGF","TGFA", 
           #regeneration 
           "MMP7", "PGR", "ESR1","AR", "GREB1", "COL1A2", "CCNO","OLFM4","CXCL10","CXCL11") 

# select the samples (using selection vector from Fig 1 d)
counts.bc.cpm.sel_IVMC <- all_counts.bc.cpm[,sel_IVMC]

# define data to plot
data2plot.bc <- data.frame()

for (gene in genes) {
    data2plot.bc <- rbind(data2plot.bc, counts.bc.cpm.sel_IVMC[rownames(counts.bc.cpm.sel_IVMC) == gene,])
}

rownames(data2plot.bc) <- genes
colnames(data2plot.bc) <- colnames(counts.bc.cpm.sel_IVMC)

# center and scale
data2plot.bc.sc <- t(scale(t(data2plot.bc)))

heatmap_selected_genes<- Heatmap(data2plot.bc.sc, cluster_columns=FALSE, cluster_rows=FALSE, 
        heatmap_legend_param = list(legend_direction = "horizontal", 
                                    title = "cpm (centered \nand scaled)", 
                                    color_bar = "continuous"),
        column_split = c(rep(0, 6),rep(1, 6),rep(2,6),rep(3, 6),rep(4,6),rep(5, 6),rep(6,6)),
        column_title = "",
        bottom_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = timepoint_colors, 
                                                                         col = timepoint_colors),
        labels = unique(metadata.sel_IVMC$description.alt_2lb), 
        labels_rot = 90,
        labels_gp = gpar(col = heatmap_labels_colors, fontsize = 8))),
        show_column_names = FALSE)

draw(heatmap_selected_genes, heatmap_legend_side="bottom")

file_name <- paste(plot_dir,"/Figure_1_E.pdf", sep = "")
pdf(file_name,height = 16,width = 8)
    draw(heatmap_selected_genes, heatmap_legend_side="bottom")
dev.off()
```


# Fig 1 F
Volcano plot highlighting selected DEGs of organoids in post-breakdown phase of the IVMC protocol (Post-breakdown 24h) compared to organoids treated with hormones but not subjected to breakdown (NB_H; no breakdown hormones)

## compare "tp4_B_H" versus "tp4_NB_H"

## define contrast
```{r}
#| label: Fig 1 f DGE contrast
(colnames(design_all) == "tp4_B_H") - (colnames(design_all) == "tp4_NB_H")
qlf_1F<- glmQLFTest(fit_all, contrast = as.vector((colnames(design_all) == "tp4_B_H") - (colnames(design_all) == "tp4_NB_H")))
tt_1F <- as.data.frame(topTags(qlf_1F,n=nrow(counts.all.asDGEList)))
head(tt_1F)
```

## define cut offs
FDR < 0.05 & logFC > 2.0 ~ up; 
FDR < 0.05 & logFC < -2.0 ~ down
```{r}
#| label: Fig 1 f DGE define cut offs
# Create significant (sig) column
results_1F <- mutate(tt_1F, sig = case_when(
                                FDR < 0.05 & logFC > 2.0 ~ up,
                                FDR < 0.05 & logFC < -2.0 ~ down,
                                TRUE ~ notsig))
sum(results_1F$sig == "Up")
sum(results_1F$sig == "Down")

file_name.table  <- paste(table_dir,"/Figure_1_F.DEG.tab",sep = "")
write.table(results_1F, file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")
```

## define genes to label 
```{r}
#| label: Fig 1 f DGE define display genes
genes_1F_up_label <- c("HS3ST2", "CSF2", "KRT6A", "DKK1", "FGF2", "KRT13", "EDAR", "CXCL6", 
                       "THBS1", "CXCL8", "CEACAM5", "EREG", "ADAMTS9", "FUT4", "AREG", "MMP7", 
                       "IL23A", "ITGA2", "PLAUR", "PCNA")
genes_1F_down_label <- c("TBC1D3K", "TGM5", "LMO1", "SEMA5B", "COL23A1", "PNCK", "ANGPTL4", 
                         "KISS1R", "CA9", "HIF3A", "ALDOC", "INHA", "PTPRR", "CXCR4", "HILPDA", 
                         "OLFM2", "NUPR1", "PAX2", "EGLN3", "VEGFA") 
genes_1F_label <- c(genes_1F_up_label,genes_1F_down_label)

# add gene labels
results_1F <- mutate(results_1F, labels = ifelse(genes %in% genes_1F_label, genes, ""))
```

## volcano plot
```{r}
#| label: Fig 1 F volcano plot
# set up base plot
p <- ggplot(data = results_1F, aes(x = logFC, y = -log10(PValue))) +
  geom_point(aes(colour = sig)) +
    xlim(-12, +12) +
  scale_color_manual(name = "|log2FC| > 2, FDR < 0.05", values = volcano_colors) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.key = element_blank(),
    legend.position = "bottom",
    
    # increase text sizes
    axis.text = element_text(size = 18), 
    axis.title = element_text(size = 22),  
    plot.title = element_text(size = 26, hjust = 0.5), 
    legend.title = element_text(size = 20), 
    legend.text = element_text(size = 20) 
  ) +
  xlab('log2 FC') +
  ylab('-log10 p-value') +
  guides(color = guide_legend(override.aes = list(size=5))) +
  
  # add gene labels
  geom_text_repel(data = filter(results_1F, labels != ""), 
                         aes(label = labels),
                         min.segment.length = 0,
                         max.overlaps = Inf,
                         size = 7,  
                         show.legend = FALSE) +
  ggtitle('Post-breakdown 24h (IVMC vs NB_H)')

print(p)

file_name <- paste(plot_dir,"/Figure_1_F.pdf", sep = "")
pdf(file_name,height = 8,width = 10)
    p
dev.off()
```


# Fig 1 G
Biological processes enriched from genes upregulated during post-breakdown phase of the IVMC protocol. 
```{r}
#| label: Fig 1 G
# all upregulated genes
genes_1F_up <- results_1F$genes[results_1F$sig == "Up"]

gostres <- gost(query = genes_1F_up, custom_bg = results_1F$genes, organism = "hsapiens", correction_method = "fdr")
head(gostres$result)
unique(gostres$result$source)

gostres_results.BP <- gostres$result[gostres$result$source == "GO:BP",]
unique(gostres_results.BP$source)

gostres_results.BP$neg_log10Pvalue <- ( -log10(gostres_results.BP$p_value) )
gostres_results.BP <- gostres_results.BP[order(gostres_results.BP$neg_log10Pvalue, decreasing = TRUE),]

head(gostres_results.BP[,c("term_name","neg_log10Pvalue")])

file_name.table  <- paste(table_dir,"/Figure_1_G.tab",sep = "")
write.table(gostres_results.BP[,c("term_name","neg_log10Pvalue")], file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")

representative_GO_terms <- c("antimicrobial humoral response",
                             "leukocyte migration",
                             "response to chemokine",
                             "chemotaxis",
                             "response to bacterium",
                             "locomotion",
                             "G protein-coupled receptor signaling pathway",
                             "cellular response to biotic stimulus",
                             "granulocyte migration",
                             "response to external stimulus",
                             "cellular response to lipopolysaccharide",
                             "humoral immune response",
                             "cell-cell signaling",
                             "negative regulation of wound healing",
                             "skin development",
                             "fibrinolysis",
                             "inflammatory response",
                             "regulation of wound healing",
                             "response to stress",
                             "intermediate filament organization",
                             "response to oxygen-containing compound",
                             "negative regulation of sprouting angiogenesis",
                             "regulation of angiogenesis")

data2plot <- gostres_results.BP[gostres_results.BP$term_name %in% representative_GO_terms,]

data2plot$term_name <- factor(data2plot$term_name, levels = rev(data2plot$term_name))

p <- ggplot(data2plot, aes(x=neg_log10Pvalue,y=term_name)) +
  geom_point(size = 10, color = "#8B1A1A") +
  xlab('-log10 p-value') +
  xlim(0, 6) +
  ylab('') +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.y = element_text(size=22)) +
  theme(axis.text.x = element_text(size=18)) +
  theme(axis.title.x =element_text(size=18)) +
  theme(panel.border = element_blank(), axis.line = element_line())

print(p)    
    
file_name <- paste(plot_dir,"/Figure_1_G.pdf", sep = "")
pdf(file_name,height = 8,width = 12)
    print(p)
dev.off() 
```


# Fig 3 C
Line plots depicting cpm of WNT7A in organoids undergoing the IVMC protocol (solid lines) and organoids not subjected to breakdown (NB_H) (dotted lines), analysed with bulk RNAseq. 

## using B0xx_tp0_NB_NH, B0xx_tp1_NB_H, B0xx_tp2_NB_H, B0xx_tp3_NB_H, B0xx_tp4_B_H, B0xx_tp5_B_H, B0xx_tp6_B_H this has been "sel_IVMC"
## plus  B0xx_tp3_NB_H, B0xx_tp4_NB_H, B0xx_tp5_NB_H, B0xx_tp6_NB_H  =>  "sel_IVMC_NB"

```{r}
#| label: Fig 3 C selection
# make a new selection (the ‘no passaging’)
metadata.sel_IVMC_NB <- mutate(as.data.frame(metadata), sel_IVMC_NB = case_when(
                                timepoint == "tp3" & hormones == "H" ~ TRUE,
                                timepoint == "tp4" & hormones == "H" & breakdown == "NB" ~ TRUE,
                                timepoint == "tp5" & hormones == "H" & breakdown == "NB" ~ TRUE,
                                timepoint == "tp6" & hormones == "H" & breakdown == "NB" ~ TRUE,
                                TRUE ~ FALSE))
    
sel_IVMC_NB <- metadata$sample[metadata.sel_IVMC_NB$sel_IVMC_NB]
sel_IVMC_NB
    
metadata.sel_IVMC_NB <- metadata[metadata.sel_IVMC_NB$sel_IVMC_NB == TRUE,]

# batch corrected data
counts.bc.cpm.sel_IVMC_NB <- all_counts.bc.cpm[,sel_IVMC_NB]
```


```{r}
#| label: Fig 3 C
metadata.sel_IVMC$general_timepoint <- factor(metadata.sel_IVMC$description.alt2, levels = (unique(metadata.sel_IVMC$description.alt2))) 
metadata.sel_IVMC_NB$general_timepoint <- factor(metadata.sel_IVMC_NB$description.alt2, levels = (unique(metadata.sel_IVMC_NB$description.alt2)))

genes_for_line_plots <- c("WNT7A")
 
for (gene in genes_for_line_plots ) {
    expr.values.sel_IVMC <- counts.bc.cpm.sel_IVMC[gene,]
    expr.values.sel_IVMC_NB <- counts.bc.cpm.sel_IVMC_NB[gene,]

    # data to plot
    df2plot <- data.frame(samples=colnames(counts.bc.cpm.sel_IVMC), cpm=expr.values.sel_IVMC)
    df2plot.sel_IVMC_NB <- data.frame(samples=colnames(counts.bc.cpm.sel_IVMC_NB), cpm=expr.values.sel_IVMC_NB)
    # combine with metadata
    df2plot_withMetadata <- merge(df2plot, metadata.sel_IVMC, by.x = "samples", by.y = "sample", sort = FALSE)
    df2plot_withMetadata.sel_IVMC_NB <- merge(df2plot.sel_IVMC_NB, metadata.sel_IVMC_NB, by.x = "samples", by.y = "sample", sort = FALSE)

    p <- ggplot(df2plot_withMetadata, aes(y=cpm)) +
          geom_line(aes(x=general_timepoint, color = donor, group = donor), linewidth=1.5, data = df2plot_withMetadata) +
          geom_line(aes(y=cpm, x=general_timepoint, color = donor, group = donor), linetype = 2, linewidth=1.5, data = df2plot_withMetadata.sel_IVMC_NB)+
          scale_color_manual(values = donor_colors)  +
          ylab(bquote(italic(.(gene))~' cpm')) +
          xlab("") +
          theme_bw() +
          theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
          theme(axis.line = element_line(colour = "black")) +
          theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95)) +
          theme(axis.text.x = element_text(size = 24)) + 
          theme(axis.text.y = element_text(size = 22)) +
          theme(axis.title.y = element_text(size = 24)) +
          theme(panel.border = element_blank()) + 
          theme(legend.text=element_text(size=22)) +
          theme(legend.title=element_text(size=22)) 
    print(p)

    file_name <- paste(plot_dir,"/Figure_3_C__",gene,".pdf", sep = "")
    pdf(file_name, height = 8,width = 10)
        print(p)
    dev.off()
}
```


# Fig 3 G
Bar plot showing the average transcripts per million (TPM) of WNT ligands at Post-breakdown 24h in EOs undergoing the IVMC protocol, analysed with bulk RNAseq.

## using TPM for B0xx_tp4_H_B
```{r}
#| label: Fig 3 G
metadata.sel3 <- mutate(as.data.frame(metadata), sel3 = case_when(
                                timepoint == "tp4" & hormones == "H" & breakdown == "B" ~ TRUE,
                                TRUE ~ FALSE))

sel3 <- metadata$sample[metadata.sel3$sel3]
sel3

tpm.selected <- all_tpm[,sel3]
    
paralogues.WNT <- rownames(tpm.selected)[grep("WNT",rownames(tpm.selected))]
# exclude ncRNA and order    
paralogues.WNT <- paralogues.WNT[c(10,5,11,6,14,17,8,7,9,15,19,2,3,13,16,12,18,4,1)]  
paralogues.WNT
    
tpm.selected.WNT <- data.frame(WNT = as.vector(as.matrix(tpm.selected[paralogues.WNT,])), 
                                paralogue = rep(paralogues.WNT,6), 
                                donor = (c(rep("B032",19), rep("B044",19), rep("B050",19), rep("B066",19), rep("B078",19), rep("B080",19))))
tpm.selected.WNT$paralogue <- factor(tpm.selected.WNT$paralogue, levels = tpm.selected.WNT$paralogue[1:19])

# and plot  
ggplot(tpm.selected.WNT, aes(y=WNT)) + 
          geom_bar(aes(x=paralogue, fill = donor), data = tpm.selected.WNT, stat="identity", position = "dodge") + 
          ylab('TPM') +
          ggtitle("Post-breakdown 24h") +
          scale_fill_manual(values = donor_colors) +
          theme_bw() +
          theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
          theme(axis.line = element_line(colour = "black")) + 
          theme(axis.text.x = element_text(angle = 90)) +
          theme(panel.border = element_blank()) 


    
p <- ggplot(tpm.selected.WNT, aes(x=paralogue,y=WNT)) + 
          stat_summary(fun=mean, geom="bar", fill="orange", ) +
          geom_point(size = 4) + 
          labs(y='TPM',x=NULL) +
          ggtitle("Post-breakdown 24h") +
          theme_bw() +
          theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
          theme(axis.line = element_line(colour = "black")) + 
          theme(axis.text.x = element_text(angle = 90, size=24, hjust=0.90,vjust=0.5)) +
          theme(axis.text.y = element_text(size=18)) +
          theme(axis.title.y =element_text(size=18)) +
          theme(axis.ticks = element_blank()) +
          theme(panel.border = element_blank()) 
    
print(p)    
    
file_name <- paste(plot_dir,"/Figure_3_G.pdf", sep = "")
pdf(file_name,height = 6,width = 8)
    print(p)
dev.off()    
```


# Fig 5 C
Heatmap depicting centered and scaled counts per million (cpm) of ligands involved in interactions between luminal early epithelial cells and other cell types, across batch corrected samples of EOs in the IVMC protocol. 

```{r}
genes <- c("WNT7A", "IL6", "CXCL1", "CXCL2", "CXCL3", "CXCL6", "CXCL8", "CXCL16", "ANXA1", "EDN1", "PDGFA", "PDGFB", 
           "AREG", "C3", "CSF1", "CSF2", "CSF3", "NAMPT", "VEGFA", "TNF", "LIF", "ADM", "GDF15", "SEMA3A","TNFSF10")

# define data to plot
data2plot.bc <- data.frame()

for (gene in genes) {
    # use counts.bc.cpm.sel_IVMC from figure 1E
    data2plot.bc <- rbind(data2plot.bc, counts.bc.cpm.sel_IVMC[rownames(counts.bc.cpm.sel_IVMC) == gene,])
}

rownames(data2plot.bc) <- genes
colnames(data2plot.bc) <- colnames(counts.bc.cpm.sel_IVMC)

# center and scale
data2plot.bc.sc <- t(scale(t(data2plot.bc)))

heatmap_selected_genes<- Heatmap(data2plot.bc.sc, cluster_columns=FALSE, cluster_rows=FALSE,
        heatmap_legend_param = list(legend_direction = "horizontal", 
                                    title = "cpm (centered \nand scaled)", 
                                    color_bar = "continuous"),
        column_split = c(rep(0, 6),rep(1, 6),rep(2,6),rep(3, 6),rep(4,6),rep(5, 6),rep(6,6)),
        column_title = "",
        bottom_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = timepoint_colors, 
                                                                         col = timepoint_colors),
        labels = unique(metadata.sel_IVMC$description.alt_2lb), 
        labels_rot = 90,
        labels_gp = gpar(col = heatmap_labels_colors, fontsize = 8))),
        show_column_names = FALSE)

draw(heatmap_selected_genes, heatmap_legend_side="bottom")

file_name <- paste(plot_dir,"/Figure_5_C.pdf", sep = "")
pdf(file_name,height = 10,width = 8)
    draw(heatmap_selected_genes, heatmap_legend_side="bottom")
dev.off()
```


# Supplementary Fig 1 C
PCA of all organoid samples from all conditions colored by the breakdown variable. 
```{r}
#| label: S Fig 1 C
colnames(metadata)[c(4,5,11)] <- c("Breakdown","Hormones","Timeline")
all_counts.CBSeq_byD.log.cpm.pca <- pca(cpm(all_counts.CBSeq_byD, log=TRUE), 
                                        metadata = metadata, removeVar = 0.1)

xlab = paste("PC1, ",round(all_counts.CBSeq_byD.log.cpm.pca$variance["PC1"],digits = 2),"% variance",sep = "")
ylab = paste("PC2, ",round(all_counts.CBSeq_byD.log.cpm.pca$variance["PC2"],digits = 2),"% variance",sep = "")

plot_object<- biplot(all_counts.CBSeq_byD.log.cpm.pca, 
       colby = 'Breakdown',
       colkey = breakdown_colors,
       shape = 'Timeline',
       pointSize = 7,
       lab = NULL,
       legendPosition = 'bottom',
       gridlines.major = FALSE,
       gridlines.minor = FALSE,
       xlab = xlab,
       ylab = ylab,
       title = "")

plot_object

file_name <- paste(plot_dir,"/S_Figure_1_C.pdf", sep = "")
    pdf(file_name, width = 8, height = 7.5)
plot_object
dev.off()
```


# Supplementary Fig 1 D
PCA of all organoid samples from all conditions colored by the hormonal treatment variable.
```{r}
#| label: S Fig 1 D
plot_object<- biplot(all_counts.CBSeq_byD.log.cpm.pca, 
       colby = 'Hormones',
       colkey = hormones_colors,
       shape = 'Timeline',
       pointSize = 7,
       lab = NULL,
       legendPosition = 'bottom',
       gridlines.major = FALSE,
       gridlines.minor = FALSE,
       xlab = xlab,
       ylab = ylab,
       title = "")

plot_object

file_name <- paste(plot_dir,"/S_Figure_1_D.pdf", sep = "")
    pdf(file_name, width = 8, height = 7.5)
plot_object
dev.off()

colnames(metadata)[c(4,5,11)] <- c("breakdown","hormones","timeline")
```


# Supplementary Fig 1 E
Hierarchically clustered heatmap displaying the top 50 differentially upregulated genes from pairwise comparisons of each timepoint against all timepoints.

## using B0xx_tp0_NB_NH, B0xx_tp1_NB_H, B0xx_tp2_NB_H, B0xx_tp3_NB_H, B0xx_tp4_B_H, B0xx_tp5_B_H, B0xx_tp6_B_H this has been "sel_IVMC"

### DGE (specific for each time point) using the raw data
```{r}
#| label: S Fig 1 E define groups
groups <- metadata.sel_IVMC$timepoint_B_H
donor <- metadata.sel_IVMC$donor
```
setting up DGEList
```{r}
#| label: S Fig 1 E prep DGE 
counts.sel_IVMC.asDGEList <- DGEList(counts=counts.sel_IVMC,
                                     genes=rownames(counts.sel_IVMC), 
                                     group = groups )

counts.sel_IVMC.asDGEList

# filter lowly expressed genes 
counts.cpm.sel_IVMC <- all_counts.cpm[,metadata.sel_IVMC$sample]
counts.sel_IVMC.asDGEList <- counts.sel_IVMC.asDGEList[ rowSums(counts.cpm.sel_IVMC > mincpm) >= minsamples, , keep.lib.size=FALSE] 
dim(counts.sel_IVMC.asDGEList)

# normalization
counts.sel_IVMC.asDGEList <- calcNormFactors(counts.sel_IVMC.asDGEList)   # same as normLibSizes(y)
head(counts.sel_IVMC.asDGEList$samples)

plotMDS(counts.sel_IVMC.asDGEList)
```

### setting up the design matrix
```{r}
#| label: S Fig 1 E design matrix
# use model.matrix
design <- model.matrix(~donor +group, data = counts.sel_IVMC.asDGEList$samples)
colnames(design)[2:12] <- substring(colnames(design)[2:12],6,14)
design
```

### dispersion estimation and fitting
```{r}
#| label: S Fig 1 E dispersion estimation and fitting
# dispersion estimation
counts.sel_IVMC.asDGEList <- estimateDisp(counts.sel_IVMC.asDGEList, design)
counts.sel_IVMC.asDGEList$common.dispersion

plotBCV(counts.sel_IVMC.asDGEList)

# fitting
fit <- glmQLFit(counts.sel_IVMC.asDGEList, design)

plotQLDisp(fit)
```

### 0) "Pre-diff" (tp0_NB_NH) specific
```{r}
#| label: S Fig 1 E DGE tp0
contrast_0 <- 1/6 * c(- (colnames(design) == "tp1_NB_H") -  (colnames(design) == "tp2_NB_H") - (colnames(design) == "tp3_NB_H") 
                      - (colnames(design) == "tp4_B_H") - (colnames(design) == "tp5_B_H") - (colnames(design) == "tp6_B_H") )
contrast_0
qlf_0 <- glmQLFTest(fit, contrast = contrast_0)
tt_0 <- as.data.frame(topTags(qlf_0,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_0)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp0_TopTable.tab",sep = "")
write.table(tt_0, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 1) "P4-diff" (tp1_NB_H) specific
```{r}
#| label: S Fig 1 E DGE tp1
contrast_1 <- c((colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp2_NB_H") - 1/6* (colnames(design) == "tp3_NB_H") 
                                - 1/6* (colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp5_B_H") - 1/6* (colnames(design) == "tp6_B_H") )
contrast_1
qlf_1 <- glmQLFTest(fit, contrast = contrast_1)
tt_1 <- as.data.frame(topTags(qlf_1,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_1)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp1_TopTable.tab",sep = "")
write.table(tt_1, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 2) "Horm Withdr 24h" (tp2_NB_H) specific
```{r}
#| label: S Fig 1 E DGE tp2
contrast_2 <-  c((colnames(design) == "tp2_NB_H") - 1/6* (colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp3_NB_H") 
                                - 1/6* (colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp5_B_H") - 1/6* (colnames(design) == "tp6_B_H") )
contrast_2
qlf_2 <- glmQLFTest(fit, contrast = contrast_2)
tt_2 <- as.data.frame(topTags(qlf_2,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_2)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp2_TopTable.tab",sep = "")
write.table(tt_2, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 3) "Horm Withdr 48h" (tp3_NB_H) specific
```{r}
#| label: S Fig 1 E DGE tp3
contrast_3 <-  c((colnames(design) == "tp3_NB_H") - 1/6* (colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp2_NB_H") 
                                - 1/6* (colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp5_B_H") - 1/6* (colnames(design) == "tp6_B_H") )
contrast_3
qlf_3 <- glmQLFTest(fit, contrast = contrast_3)
tt_3 <- as.data.frame(topTags(qlf_3,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_3)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp3_TopTable.tab",sep = "")
write.table(tt_3, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 4) "Post-breakdown 24h" (tp4_B_H) specific
```{r}
#| label: S Fig 1 E DGE tp4
contrast_4 <-  c((colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp2_NB_H") - 1/6* (colnames(design) == "tp3_NB_H") 
                                - 1/6* (colnames(design) == "tp5_B_H") - 1/6* (colnames(design) == "tp6_B_H") )
contrast_4
qlf_4 <- glmQLFTest(fit, contrast = contrast_4)
tt_4 <- as.data.frame(topTags(qlf_4,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_4)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp4_TopTable.tab",sep = "")
write.table(tt_4, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 5) "E2-diff 24h" (tp5_B_H) specific
```{r}
#| label: S Fig 1 E DGE tp5
contrast_5 <- c((colnames(design) == "tp5_B_H") - 1/6* (colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp2_NB_H") - 1/6* (colnames(design) == "tp3_NB_H") 
                                - 1/6* (colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp6_B_H") )
qlf_5 <- glmQLFTest(fit, contrast = contrast_5)
contrast_5
tt_5 <- as.data.frame(topTags(qlf_5,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_5)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp5_TopTable.tab",sep = "")
write.table(tt_5, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

### 6) "E2 diff 48h" (tp6_B_H) specific
```{r}
#| label: S Fig 1 E DGE tp6
contrast_6 <-  c((colnames(design) == "tp6_B_H") - 1/6* (colnames(design) == "tp1_NB_H") - 1/6* (colnames(design) == "tp2_NB_H") - 1/6* (colnames(design) == "tp3_NB_H") 
                                - 1/6* (colnames(design) == "tp4_B_H") - 1/6* (colnames(design) == "tp5_B_H") )
qlf_6 <- glmQLFTest(fit, contrast = contrast_6)
contrast_6
tt_6 <- as.data.frame(topTags(qlf_6,n=nrow(counts.sel_IVMC.asDGEList)))
head(tt_6)

file_name.table <- paste(table_dir,"/S_Figure_1_E.tp6_TopTable.tab",sep = "")
write.table(tt_6, file_name.table, quote = FALSE, row.names = FALSE, sep = "\t")
```

## gene list Heatmaps
```{r}
#| label: S Fig 1 E heatmap
# top 50 genes with logFC > 2
genes_top50up_all <- c(tt_0[tt_0$logFC > 2,]$genes[1:50],
                       tt_1[tt_1$logFC > 2,]$genes[1:50],
                       tt_2[tt_2$logFC > 2,]$genes[1:50],
                       tt_3[tt_3$logFC > 2,]$genes[1:50],
                       tt_4[tt_4$logFC > 2,]$genes[1:50],
                       tt_5[tt_5$logFC > 2,]$genes[1:50],
                       tt_6[tt_6$logFC > 2,]$genes[1:50])

# define data to plot
data2plot <- data.frame()

for (gene in genes_top50up_all) {
    data2plot <- rbind(data2plot, counts.bc.cpm.sel_IVMC[rownames(counts.bc.cpm.sel_IVMC) == gene,])
}
# rownames(data2plot) <- genes
colnames(data2plot) <- colnames(counts.bc.cpm.sel_IVMC)

# center and scale
data2plot.sc <- t(scale(t(data2plot)))

h <- Heatmap(data2plot.sc, cluster_columns=FALSE, cluster_rows=FALSE, 
        heatmap_legend_param = list(title = "cpm\n(centered\nscaled)", color_bar = "continuous"),
        column_split = c(rep(0, 6),rep(1, 6),rep(2,6),rep(3, 6),rep(4,6),rep(5, 6),rep(6,6)),
        row_split = c(rep(0,50),rep(1,50),rep(2,50),rep(3,50),rep(4,50),rep(5,50),rep(6,50)),
        column_title = " ",
        bottom_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = timepoint_colors, 
                                                                         col = timepoint_colors),
                                                               labels = unique(metadata.sel_IVMC$description.alt_2lb), 
                                                               labels_rot = 90,
                                                               labels_gp = gpar(col = heatmap_labels_colors, 
                                                                                fontsize = 6))),
        show_column_names = FALSE,
        row_title = NULL)

print(h)

file_name <- paste(plot_dir,"/S_Figure_1_E.pdf",sep = "")
pdf(file_name, width = 6, height = 7)    
    print(h)
dev.off()
```


# Supplementary Fig 1 F
Heatmap depicting batch corrected centered and scaled counts per million (cpm) of selected genes across the IVMC protocol (comparing hormone treatmenT)
```{r}
genes <- c("PDK1","SLC2A1","HK2","ALDOA","GAPDH","ENO2","SERPINE1", 
           "VEGFA","FLT1","ANGPT4","ANGPTL4","ANGPTL6","ANGPTL8","ANG","IGF1","PTGS2","TIMP1","EGF","PDHA1",  
           "CXCL8","FGF2","PDGFB","EFNB2","TFRC","HMOX1","EDN1","NOS1","NOS2","NOS3","CDKN1A") 
# selection
selm <- c(metadata$sample[metadata$timepoint == "tp0"],
          metadata$sample[metadata$timepoint == "tp1" & metadata$hormones == "NH"],
          metadata$sample[metadata$timepoint == "tp1" & metadata$hormones == "H"],
          metadata$sample[metadata$timepoint == "tp2" & metadata$hormones == "NH"],
          metadata$sample[metadata$timepoint == "tp2" & metadata$hormones == "H"],          
          metadata$sample[metadata$timepoint == "tp3" & metadata$hormones == "NH"],
          metadata$sample[metadata$timepoint == "tp3" & metadata$hormones == "H"],     
          metadata$sample[metadata$timepoint == "tp4" & metadata$hormones == "NH" & metadata$breakdown == "B"],
          metadata$sample[metadata$timepoint == "tp4" & metadata$hormones == "H" & metadata$breakdown == "B"],          
          metadata$sample[metadata$timepoint == "tp5" & metadata$hormones == "NH" & metadata$breakdown == "B"],
          metadata$sample[metadata$timepoint == "tp5" & metadata$hormones == "H" & metadata$breakdown == "B"],  
          metadata$sample[metadata$timepoint == "tp6" & metadata$hormones == "NH" & metadata$breakdown == "B"],
          metadata$sample[metadata$timepoint == "tp6" & metadata$hormones == "H" & metadata$breakdown == "B"])

selm

# select the samples 
counts.bc.cpm.selm <- all_counts.bc.cpm[,selm]

metadata.selm <- metadata[metadata$sample %in% selm,]

# define data to plot
data2plot.bc <- data.frame()

for (gene in genes) {
    data2plot.bc <- rbind(data2plot.bc, counts.bc.cpm.selm[rownames(counts.bc.cpm.selm) == gene,])
}

rownames(data2plot.bc) <- genes
colnames(data2plot.bc) <- colnames(counts.bc.cpm.selm)

# center and scale
data2plot.bc.sc <- t(scale(t(data2plot.bc)))

# heatmap with annotation (trying with two columns)
heatmap_selected_genes <- Heatmap(data2plot.bc.sc, cluster_columns=FALSE, cluster_rows=FALSE, 
        heatmap_legend_param = list(legend_direction = "horizontal", 
                                    title = "cpm (centered \nand scaled)", 
                                    color_bar = "continuous"),
        column_split = c(rep(0, 6),rep(1,6),rep(2,6),rep(3,6),rep(4,6),rep(5,6),rep(6,6),rep(7,6),rep(8,6),rep(9,6),rep(10,6),rep(11,6),rep(12,6)),
        column_title = "",
        bottom_annotation = HeatmapAnnotation(row1 = anno_block(gp = gpar(fill = c("#CCCCC4",rep(c("#CCCCC4","#00C5CD"),6)), 
                                                                          col = c("#CCCCC4",rep(c("#CCCCC4","#00C5CD"),6))),
                                                                labels = c("NH",rep(c("NH","H"),6)), 
                                                                labels_gp = gpar(col = "black", fontsize = 8)),
                                              row2 = anno_block(gp = gpar(fill = c("#e6e6e6",rep("#5e17d1",2),rep("#989692",4),rep("#cb3b63",2),rep("#debaf5",2),rep("#BE9FD2",2)), 
                                                                         col = c("#e6e6e6",rep("#5e17d1",2),rep("#989692",4),rep("#cb3b63",2),rep("#debaf5",2),rep("#BE9FD2",2))),
                                                                labels_rot = 90,
                                                               #labels = unique(metadata.sel_IVMC$description.alt_2lb), 
                                                               labels = c("Pre-diff",rep("P4-diff",2),rep("Horm\nWithdr\n24h",2),rep("Horm\nWithdr\n48h",2),
                                                                          rep("Post-\nbreakdown\n24h",2),rep("E2-diff\n24h",2),rep("E2-diff\n48h",2)), 
                                                               labels_gp = gpar(col = c("black",rep("white",2),rep("black",4),rep("white",2),rep("black",4)), fontsize = 8))
                                              ),
        show_column_names = FALSE)

draw(heatmap_selected_genes, heatmap_legend_side="bottom")

file_name <- paste(plot_dir,"/S_Figure_1_F.pdf", sep = "")
pdf(file_name,height = 10,width = 8)
    draw(heatmap_selected_genes, heatmap_legend_side="bottom")
dev.off()
```

# Supplementary Fig 2 A

## using B0xx_tp0_NB_NH, B0xx_tp1_NB_H, B0xx_tp2_NB_H, B0xx_tp3_NB_H, B0xx_tp4_B_H, B0xx_tp5_B_H, B0xx_tp6_B_H this has been "sel_IVMC"
Line plots showing the (batch coreccted) mean expression levels (in cpm) of genes reported in literature to mark stem cells of the endometrial epithelium (MKI67, ALDH1A1, CDH2, SOX9, FUT4) across different timepoints in the IVMC protocol. The red line represents the mean cpm values across donors, while the shaded blue region indicates the standard deviation around the mean. 
```{r}
#| label: S Fig 2 A
genes_for_line_plots <- c("MKI67","ALDH1A1","SOX9", "FUT4")

for (gene in genes_for_line_plots ) {
    expr.values.sel_IVMC <- counts.bc.cpm.sel_IVMC[gene,]
    
    # data to plot
    df2plot <- data.frame(samples=colnames(counts.bc.cpm.sel_IVMC), cpm=expr.values.sel_IVMC)
    
    # combine with metadata
    df2plot_withMetadata <- merge(df2plot, metadata.sel_IVMC, by.x = "samples", by.y = "sample", sort = FALSE)
    
    # aggregate
    df2plot_aggregate <- data.frame(general_timepoint = aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=mean)[,1],
                                    cpm_mean = aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=mean)[,2],
                                    cpm_mean_plus_sd = aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=mean)[,2] +
                                        aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=sd)[,2],
                                    cpm_mean_minus_sd = aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=mean)[,2] -
                                        aggregate(df2plot_withMetadata$cpm, list(df2plot_withMetadata$general_timepoint), FUN=sd)[,2] )
    
    
    p <- ggplot(df2plot_aggregate, aes(y=cpm_mean, x=general_timepoint, group = 1)) + 
        geom_line(color = "red") +
        geom_point(color = "red", size = 2) +
        geom_ribbon(aes(x = general_timepoint, ymax = cpm_mean_plus_sd, ymin = cpm_mean_minus_sd), 
                    alpha = 0.3, fill = "lightblue") +
        ggtitle(bquote(italic(.(gene)))) +
        ylab("cpm (Mean ± SD across donors)") +
        xlab("") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
        theme(axis.line = element_line(colour = "black")) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 24)) +
        theme(axis.text.y = element_text(size = 22)) +
        theme(axis.title.y = element_text(size = 24)) +
        theme(panel.border = element_blank())
    print(p)
    
    file_name <- paste(plot_dir,"/S_Figure_2_A__",gene,".pdf", sep = "")
    pdf(file_name, height = 8,width = 10)
    print(p)
    dev.off()
}
```


# Supplementary Fig 3 A
Volcano plot highlighting selected from comparison between organoids at P4-diff phase of the IVMC protocol to organoids not treated with hormones but undergone breakdown (B_NH) at the same timepoint. 

## define contrast
```{r}
#| label: S Fig 3 A DGE contrast
(colnames(design_all) == "tp1_NB_H") - (colnames(design_all) == "tp1_NB_NH")
qlf_3A<- glmQLFTest(fit_all, contrast = as.vector((colnames(design_all) == "tp1_NB_H") - (colnames(design_all) == "tp1_NB_NH")))
tt_3A <- as.data.frame(topTags(qlf_3A,n=nrow(counts.all.asDGEList)))
head(tt_3A)
```

## define cut offs
FDR < 0.05 & logFC > 2.0 ~ up; 
FDR < 0.05 & logFC < -2.0 ~ down
```{r}
#| label: S Fig 3 A DGE define cut offs
# create significant (sig) column
results_3A <- mutate(tt_3A, sig = case_when(
    FDR < 0.05 & logFC > 2.0 ~ up,
    FDR < 0.05 & logFC < -2.0 ~ down,
    TRUE ~ notsig))
sum(results_3A$sig == "Up")
sum(results_3A$sig == "Down")

file_name.table  <- paste(table_dir,"/S_Figure_3_A.DEG.tab",sep = "")
write.table(results_3A, file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")
```

## define genes to label 
```{r}
#| label: S Fig 3 A DGE define display genes
genes_3A_up_label <- c("SULT1E1", "GATA2", "SCGB1D4", "ALPG", "PAEP", "SCGB1D2", 
                       "SERPINA5", "S100P", "SCGB2A2", "SCGB2A1", "PTGS2", "ENPP3", 
                       "KRT17", "TPPP3", "IL1A")
genes_3A_down_label <- c("NCAM1", "SHISA3", "KRT4", "FZD10", "SSTR1", "SCTR", 
                         "FAM20A", "PTPRG", "RXRG", "MMP10", "FN1", "S100A9", 
                         "MMP3", "MMP1", "ASCL2")
genes_3A_label <- c(genes_3A_up_label,genes_3A_down_label)

# add gene labels
results_3A <- mutate(results_3A, labels = ifelse(genes %in% genes_3A_label, genes, ""))
```

## volcano plot
```{r}
#| label: S Fig 3 A volcano plot
# set up base plot
p <- ggplot(data = results_3A, aes(x = logFC, y = -log10(PValue))) +
    geom_point(aes(colour = sig)) +
    xlim(-12, +12) +
    scale_color_manual(name = "|log2FC| > 2, FDR < 0.05", values = volcano_colors) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank(),
        legend.position = "bottom",
        
        # increase text sizes
        axis.text = element_text(size = 18), 
        axis.title = element_text(size = 22),  
        plot.title = element_text(size = 26, hjust = 0.5), 
        legend.title = element_text(size = 20), 
        legend.text = element_text(size = 20) 
    ) +
    xlab('log2 FC') +
    ylab('-log10 p-value') +
    guides(color = guide_legend(override.aes = list(size=5))) +
    
    # add gene labels
    geom_text_repel(data = filter(results_3A, labels != ""), 
                    aes(label = labels),
                    min.segment.length = 0,
                    max.overlaps = Inf,
                    size = 7,  
                    show.legend = FALSE) +
    ggtitle('P4-diff (IVMC vs NH)')

print(p)

file_name <- paste(plot_dir,"/S_Figure_3_A.pdf", sep = "")
pdf(file_name,height = 8,width = 10)
p
dev.off()
```


# Supplementary Fig 3 C
Volcano plot highlighting selected DEGs from comparison between organoids of the IVMC protocol at P4-diff phase to organoids at hormonal withdrawal 48 h. 

## define contrast
```{r}
#| label: S Fig 3 C DGE contrast
(colnames(design_all) == "tp3_NB_H") - (colnames(design_all) == "tp1_NB_H")  - ((colnames(design_all) == "tp3_NB_NH") - (colnames(design_all) == "tp1_NB_NH") )
qlf_3C<- glmQLFTest(fit_all, contrast = as.vector( (colnames(design_all) == "tp3_NB_H") - (colnames(design_all) == "tp1_NB_H")  - ((colnames(design_all) == "tp3_NB_NH") - (colnames(design_all) == "tp1_NB_NH") ) ) )
tt_3C <- as.data.frame(topTags(qlf_3C,n=nrow(counts.all.asDGEList)))
head(tt_3C)
```

## define cut offs
FDR < 0.05 & logFC > 2.0 ~ up; 
FDR < 0.05 & logFC < -2.0 ~ down
```{r}
#| label: S Fig 3 C DGE define cut offs
# create significant (sig) column
results_3C <- mutate(tt_3C, sig = case_when(
    FDR < 0.05 & logFC > 2.0 ~ up,
    FDR < 0.05 & logFC < -2.0 ~ down,
    TRUE ~ notsig))
sum(results_3C$sig == "Up")
sum(results_3C$sig == "Down")

file_name.table  <- paste(table_dir,"/S_Figure_3_C.DEG.tab",sep = "")
write.table(results_3C, file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")
```

## define genes to label 
```{r}
#| label: S Fig 3 C DGE define display genes
genes_3C_up_label <- c("HMGA2-AS1", "SIGLEC15", "MEOX1", "VTCN1", "RXRG", "MT1E", 
                       "MMP10", "CASP1", "MT1G", "SSTR1")
genes_3C_down_label <- c("PHACTR3", "KRT16", "SERPINA3", "KRT17", "SFRP1", "NOTUM", 
                         "S100P", "CXCL8", "PAEP", "PLAUR", "AREG")
genes_3C_label <- c(genes_3C_up_label,genes_3C_down_label)

# add gene labels
results_3C <- mutate(results_3C, labels = ifelse(genes %in% genes_3C_label, genes, ""))
```

## volcano plot
```{r}
#| label: S Fig 3 C volcano plot
# set up base plot
p <- ggplot(data = results_3C, aes(x = logFC, y = -log10(PValue))) +
    geom_point(aes(colour = sig)) +
    xlim(-12, +12) +
    scale_color_manual(name = "|log2FC| > 2, FDR < 0.05", values = volcano_colors) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank(),
        legend.position = "bottom",
        
        # increase text sizes
        axis.text = element_text(size = 18), 
        axis.title = element_text(size = 22),  
        plot.title = element_text(size = 26, hjust = 0.5), 
        legend.title = element_text(size = 20), 
        legend.text = element_text(size = 20) 
    ) +
    xlab('log2 FC') +
    ylab('-log10 p-value') +
    guides(color = guide_legend(override.aes = list(size=5))) +
    
    # add gene labels 4
    geom_text_repel(data = filter(results_3C, labels != ""), 
                    aes(label = labels),
                    min.segment.length = 0,
                    max.overlaps = Inf,
                    size = 7,  
                    show.legend = FALSE) +
    ggtitle('Horm.Withdr. 48h vs P4-diff (controlled for hormone)')

print(p)

file_name <- paste(plot_dir,"/S_Figure_3_C.pdf", sep = "")
pdf(file_name,height = 8,width = 10)
p
dev.off()
```


# Supplementary Fig 3 F
Volcano plot highlighting selected DEGs comparing EOs at Post-breakdown 24h phase of the IVMC protocol to control EOs (B_NH) at the same timepoint. 

```{r}
#| label: S Fig 3 F DGE contrast
(colnames(design_all) == "tp4_B_H") - (colnames(design_all) == "tp4_B_NH")
qlf_3F<- glmQLFTest(fit_all, contrast = as.vector((colnames(design_all) == "tp4_B_H") - (colnames(design_all) == "tp4_B_NH")))
tt_3F <- as.data.frame(topTags(qlf_3F,n=nrow(counts.all.asDGEList)))
head(tt_3F)
```

## define cut offs
FDR < 0.05 & logFC > 2.0 ~ up; 
FDR < 0.05 & logFC < -2.0 ~ down
```{r}
#| label: S Fig 3 F DGE define cut offs
# create significant (sig) column
results_3F <- mutate(tt_3F, sig = case_when(
    FDR < 0.05 & logFC > 2.0 ~ up,
    FDR < 0.05 & logFC < -2.0 ~ down,
    TRUE ~ notsig))
sum(results_3F$sig == "Up")
sum(results_3F$sig == "Down")

file_name.table  <- paste(table_dir,"/S_Figure_3_F.DEG.tab",sep = "")
write.table(results_3F, file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")
```

## define genes to label 
```{r}
#| label: S Fig 3 F DGE define display genes
genes_3F_up_label <- c("DUOXA2", "DUOX2", " IL19", "SAA2", "SAA1", "SAA4", "TREM1", 
                       "CCR7", "MUC13", "IL10", "CCL20", "IRAK3", "PF4V1")
genes_3F_down_label <- c("FBN2", "CALCA", "SHISA3", "KRT4", "CYP1A2", "NGFR", "KRT5",
                         "ESRRB", "SSTR1", "PLXNA4", "FRZB")
genes_3F_label <- c(genes_3F_up_label,genes_3F_down_label)

# add gene labels
results_3F <- mutate(results_3F, labels = ifelse(genes %in% genes_3F_label, genes, ""))
```

## volcano plot
```{r}
#| label: S Fig 3 F volcano plot
# set up base plot
p <- ggplot(data = results_3F, aes(x = logFC, y = -log10(PValue))) +
    geom_point(aes(colour = sig)) +
    xlim(-12, +12) +
    scale_color_manual(name = "|log2FC| > 2, FDR < 0.05", values = volcano_colors) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank(),
        legend.position = "bottom",
        
        # increase text sizes
        axis.text = element_text(size = 18), 
        axis.title = element_text(size = 22),  
        plot.title = element_text(size = 26, hjust = 0.5), 
        legend.title = element_text(size = 20), 
        legend.text = element_text(size = 20) 
    ) +
    xlab('log2 FC') +
    ylab('-log10 p-value') +
    guides(color = guide_legend(override.aes = list(size=5))) +
    
    # add gene labels 
    geom_text_repel(data = filter(results_3F, labels != ""), 
                    aes(label = labels),
                    min.segment.length = 0,
                    max.overlaps = Inf,
                    size = 7,  
                    show.legend = FALSE) +
    ggtitle('Post-breakdown 24h (IVMC vs B_NH)')

print(p)

file_name <- paste(plot_dir,"/S_Figure_3_F.pdf", sep = "")
pdf(file_name,height = 8,width = 10)
p
dev.off()
```


# Supplementary Fig 3 G
Biological processes enriched in Post-breakdown 24h phase of the IVMC protocol using genes upregulated in A. 

```{r}
#| label: S Fig 3G
# all upregulated genes
genes_3F_up <- results_3F$genes[results_3F$sig == "Up"]

gostres <- gost(query = genes_3F_up, custom_bg = results_3F$genes, organism = "hsapiens", correction_method = "fdr")
head(gostres$result)
unique(gostres$result$source)

gostres_results.BP <- gostres$result[gostres$result$source == "GO:BP",]
unique(gostres_results.BP$source)

gostres_results.BP$neg_log10Pvalue <- ( -log10(gostres_results.BP$p_value) )
gostres_results.BP <- gostres_results.BP[order(gostres_results.BP$neg_log10Pvalue, decreasing = TRUE),]
head(gostres_results.BP[,c("term_name","neg_log10Pvalue")])

file_name.table  <- paste(table_dir,"/S_Figure_3_G.tab",sep = "")
write.table(gostres_results.BP[,c("term_name","neg_log10Pvalue")], file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")

representative_GO_terms <- c("cilium movement", 
                             "cell motility", 
                             "acute inflammatory response", 
                             "acute-phase response", 
                             "defense response",
                             "sperm motility",
                             "response to bacterium",
                             "leukocyte migration",
                             "neutrophil chemotaxis",
                             "extracellular transport")

head(gostres_results.BP[,c("term_name","neg_log10Pvalue")])
gostres_results.BP$term_name[gostres_results.BP$term_name %in% representative_GO_terms]

data2plot <- gostres_results.BP[gostres_results.BP$term_name %in% representative_GO_terms,]

data2plot$term_name <- factor(data2plot$term_name, levels = rev(data2plot$term_name))

p <- ggplot(data2plot, aes(x=neg_log10Pvalue,y=term_name)) +
    geom_point(size = 10, color = "#00C5CD") +
    xlab('-log10 p-value') +
    xlim(0, 10) +
    ylab('') +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(axis.text.y = element_text(size=22)) +
    theme(axis.text.x = element_text(size=18)) +
    theme(axis.title.x =element_text(size=18)) +
    theme(panel.border = element_blank(), axis.line = element_line())

print(p)    

file_name <- paste(plot_dir,"/S_Figure_3_G.pdf", sep = "")
pdf(file_name,height = 8,width = 7)
print(p)
dev.off() 
```


# Supplementary Fig 3 I
Volcano plot highlighting selected DEGs from comparison of organoids between E2-diff 48h phase of the IVMC protocol to organoids not treated with hormones but undergone breakdown (B_NH) at the same timepoint. 

## define contrast
```{r}
#| label: S Fig 3 I DGE contrast
(colnames(design_all) == "tp6_B_H") - (colnames(design_all) == "tp6_B_NH")
qlf_3I<- glmQLFTest(fit_all, contrast = as.vector((colnames(design_all) == "tp6_B_H") - (colnames(design_all) == "tp6_B_NH")))
tt_3I <- as.data.frame(topTags(qlf_3I,n=nrow(counts.all.asDGEList)))
head(tt_3I)
```

## define cut offs
FDR < 0.05 & logFC > 2.0 ~ up; 
FDR < 0.05 & logFC < -2.0 ~ down
```{r}
#| label: S Fig 3 I DGE define cut offs
# create significant (sig) column
results_3I <- mutate(tt_3I, sig = case_when(
    FDR < 0.05 & logFC > 2.0 ~ up,
    FDR < 0.05 & logFC < -2.0 ~ down,
    TRUE ~ notsig))
sum(results_3I$sig == "Up")
sum(results_3I$sig == "Down")

file_name.table  <- paste(table_dir,"/S_Figure_3_I.DEG.tab",sep = "")
write.table(results_3I, file_name.table,
            quote = FALSE, row.names = FALSE, sep = "\t")
```

## define genes to label 
```{r}
#| label: S Fig 3 I DGE define display genes
genes_3I_up_label <- c("OLFM4", "COL1A2", "PGR", "CDC20B", "GMNC", "GREB1", "GATA5", "CCN5")
genes_3I_down_label <- c("KRT5", "MMP1", "PAX2", "FRZB", "MMP10", "SHISA3", "PLXNA4", "SSTR1")
genes_3I_label <- c(genes_3I_up_label,genes_3I_down_label)

# add gene labels
results_3I <- mutate(results_3I, labels = ifelse(genes %in% genes_3I_label, genes, ""))
```

## volcano plot
```{r}
#| label: S Fig 3 I volcano plot
# set up base plot
p <- ggplot(data = results_3I, aes(x = logFC, y = -log10(PValue))) +
    geom_point(aes(colour = sig)) +
    xlim(-12, +12) +
    scale_color_manual(name = "|log2FC| > 2, FDR < 0.05", values = volcano_colors) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank(),
        legend.position = "bottom",
        
        # increase text sizes
        axis.text = element_text(size = 18), 
        axis.title = element_text(size = 22),  
        plot.title = element_text(size = 26, hjust = 0.5), 
        legend.title = element_text(size = 20), 
        legend.text = element_text(size = 20) 
    ) +
    xlab('log2 FC') +
    ylab('-log10 p-value') +
    guides(color = guide_legend(override.aes = list(size=5))) +
    
    # add gene labels 4
    geom_text_repel(data = filter(results_3I, labels != ""), 
                    aes(label = labels),
                    min.segment.length = 0,
                    max.overlaps = Inf,
                    size = 7,  
                    show.legend = FALSE) +
    ggtitle('E2-diff 48h (IVMC vs B_NH)')

print(p)

file_name <- paste(plot_dir,"/S_Figure_3_I.pdf", sep = "")
pdf(file_name,height = 8,width = 10)
p
dev.off()
```



# Supplementary Fig 5 F
Bar plot showing the average transcripts per million (TPM) of FZD receptors during at Post-breakdown 24h in EOs undergoing the IVMC protocol, analysed with bulk RNAseq. 

## using TPM for B0xx_tp4_H_B
```{r}
#| label: S Fig 5 F
# use tpm.selected from 10 a
paralogues.FZD <- rownames(tpm.selected)[grep("FZD",rownames(tpm.selected))]
# exclude ncRNA and order    
paralogues.FZD <- paralogues.FZD[c(4,9,1,7,5,6,3,8,10,2)]  
paralogues.FZD

tpm.selected.FZD <- data.frame(FZD = as.vector(as.matrix(tpm.selected[paralogues.FZD,])), 
                               paralogue = rep(paralogues.FZD,6), 
                               donor = (c(rep("B032",10), rep("B044",10), rep("B050",10), rep("B066",10), rep("B078",10), rep("B080",10))))
tpm.selected.FZD$paralogue <- factor(tpm.selected.FZD$paralogue, levels = tpm.selected.FZD$paralogue[1:10])

# and plot  
ggplot(tpm.selected.FZD, aes(y=FZD)) + 
    geom_bar(aes(x=paralogue, fill = donor), data = tpm.selected.FZD, stat="identity", position = "dodge") + 
    ylab('TPM') +
    ggtitle("Post-breakdown 24h") +
    scale_fill_manual(values = donor_colors) +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(axis.line = element_line(colour = "black")) + 
    theme(axis.text.x = element_text(angle = 90)) +
    theme(panel.border = element_blank()) 



p <- ggplot(tpm.selected.FZD, aes(x=paralogue,y=FZD)) + 
    stat_summary(fun=mean, geom="bar", fill="magenta4", ) +
    geom_point(size = 4) + 
    labs(y='TPM',x=NULL) +
    ggtitle("Post-breakdown 24h") +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(axis.line = element_line(colour = "black")) + 
    theme(axis.text.x = element_text(angle = 90, size=24, hjust=0.90,vjust=0.5)) +
    theme(axis.text.y = element_text(size=18)) +
    theme(axis.title.y =element_text(size=18)) +
    theme(axis.ticks = element_blank()) +
    theme(panel.border = element_blank()) 

print(p)    

file_name <- paste(plot_dir,"/S_Figure_5_F.pdf", sep = "")
pdf(file_name,height = 6,width = 8)
print(p)
dev.off()    
```


# sessionInfo
```{r}
sessionInfo()
```