---
title: "Generating an epithelial endometrial atlas"
date: "`r Sys.Date()`"
format: 
    html:
        toc: true
        theme: united
embed-resources: true
---

# Load packages

```{r load-pkgs}
suppressPackageStartupMessages({
    library(fishpond)
    library(SummarizedExperiment)
    library(SingleCellExperiment)
    library(scuttle)
    library(scater)
    library(scran)
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(msigdbr)
    library(batchelor)
    library(cowplot)
    library(ComplexHeatmap)
    library(circlize)
    library(scDblFinder)
    library(Seurat)
    library(sketchR)
    library(tricycle)
    library(org.Hs.eg.db)
    library(parallel)
    library(bluster)
    library(BiocParallel)
    library(igraph)
    library(DT)
})
```

# Define parameters and reference files

```{r define-params}
samples <- c("4861STDY7387181", "4861STDY7387182", "4861STDY7387183", 
             "4861STDY7771115", "4861STDY7771123", "MRC_Endo8625698", 
             "MRC_Endo8625699", "MRC_Endo8712024", "MRC_Endo8712032", 
             "MRC_Endo8715415", "MRC_Endo8715416", "GSM4577306", "GSM4577307", 
             "GSM4577308", "GSM4577309", "GSM4577310", "GSM4577311", 
             "GSM4577312", "GSM4577313", "GSM4577314", "GSM4577315", 
             "Shih_MEtissue1", "Shih_MEtissue2", "Shih_MEtissue3", 
             "Shih_MEtissue4", "Shih_MEtissue5", "Shih_MEtissue6", 
             "Shih_MEtissue7", "Huang_GSM6605437", "Huang_GSM6605438", 
             "Huang_GSM6605440", "Huang_GSM6605439", "Huang_GSM7277296", 
             "Huang_GSM7277297", "Huang_GSM7277298", "Lai_GSM5572238", 
             "Lai_GSM5572239", "Lai_GSM5572240")
blockvar <- "sample"           ## used as blocking factor in doublet annotation
batchvar <- "donorCol"         ## used for batch correction
datasetvar <- "matchDataset"   ## used as blocking factor in HVG annotation

## Sample annotation
sampleannot <- "invivo_atlas_samples.txt"

## Gene annotation (derived from GENCODE v38 gtf file)
t2g <- "reference/gencode.v38.annotation.tx2gene_full.rds"

## Reproductive cell atlas data (downloaded from 
## https://www.reproductivecellatlas.org/ and converted to SingleCellExperiment 
## format)
rca <- "data/reproductivecellatlas/2021/endometrium_all.rds"
heca <- "data/reproductivecellatlas/2023/endometriumAtlasV2_cells_with_counts.rds"

## Cell cycle marker genes
## https://github.com/ventolab/HECA-Human-Endometrial-Cell-Atlas/tree/main/utils
s_genes <- "data/reproductivecellatlas/2023/S_genes_HECA.tsv"
g2m_genes <- "data/reproductivecellatlas/2023/G2M_genes_HECA.tsv"

## Shih et al cell annotations (obtained from 
## https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE203191)
shih_cellannot <- "data/GSE203191/GSE203191_Shih_endo_meta_frame.tsv"
```

# Assemble gene annotation

```{r gene-annot}
geneannot <- readRDS(t2g) |>
    dplyr::select(gene_id, seqnames, gene_type, gene_name) |>
    dplyr::rename(chromosome = seqnames) |>
    dplyr::distinct()

## Annotate ribosomal protein genes
## GO term "structural constituent of ribosome" (GO:0003735)
golist <- as.list(org.Hs.egGO2ALLEGS)
ribo_entrez <- unique(golist[["GO:0003735"]])
length(ribo_entrez)
tmp <- select(org.Hs.eg.db, keys = ribo_entrez, columns = "ENSEMBL", 
              keytype = "ENTREZID")
ribo_ensembl <- unique(na.omit(tmp[, "ENSEMBL"]))
length(ribo_ensembl)
summary(ribo_ensembl %in% sub("\\.[0-9]+$", "", geneannot$gene_id))
geneannot$is_ribosomal_protein <- 
    sub("\\.[0-9]+$", "", geneannot$gene_id) %in% ribo_ensembl
geneannot$gene_type_ribo <- ifelse(
    geneannot$gene_type == "protein_coding",
    ifelse(geneannot$is_ribosomal_protein, "protein_coding_ribosome",
           geneannot$gene_type),
    geneannot$gene_type)
```

# Read data

```{r read-data}
#| message: false

## Get sample IDs
tbl <- read.delim(sampleannot) |>
     dplyr::filter(SampleName %in% samples)
DT::datatable(tbl, options = list(scrollX = TRUE, pageLength = 50))
sns <- unique(tbl$SampleName)
names(sns) <- sns

## Read alevin-fry data and combine
dat <- do.call(cbind, lapply(sns, function(s) {
    x <- fishpond::loadFry(
        fryDir = file.path("data", "alevin_fry_splici", s), 
        outputFormat = list(counts = c("S", "A"), spliced = c("S", "A"),
                            unspliced = c("U"), total = c("U", "S", "A")))
    rowData(x)$gene_name <- 
        geneannot$gene_name[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$chromosome <- 
        geneannot$chromosome[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$gene_type <- 
        geneannot$gene_type[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$gene_type_ribo <- 
        geneannot$gene_type_ribo[match(rowData(x)$gene_ids, geneannot$gene_id)]
    
    ## Keep only cells with at least 1,000 detected genes
    x <- x[, colSums(assay(x, "total") > 0) > 1000]
    
    ## Add sample ID to colnames
    x$sample <- s
    colnames(x) <- paste0(s, "_", colnames(x))
    x
}))
dat

## Set rownames to gene names where possible
rownames(dat) <- scater::uniquifyFeatureNames(
    ID = rownames(dat), 
    names = rowData(dat)$gene_name
)
```

# Add sample annotations

```{r add-sample-annots}
## Extract barcodes corresponding to control samples in Shih data and remove 
## all other barcodes from the Shih samples
cb <- read.delim(shih_cellannot) |>
    dplyr::filter(pheno == "Control") |>
    dplyr::mutate(barcode = gsub("-[0-9]$", "", barcode)) |>
    dplyr::mutate(sample_cb = paste("Shih", run, barcode, sep = "_")) |>
    dplyr::pull(sample_cb)
to_remove <- which(grepl("Shih", dat$sample) & 
                       !paste0(dat$sample, "_", dat$barcodes) %in% cb)
dat <- dat[, -to_remove]
dim(dat)

## Add 'cell' column
dat$cell <- colnames(dat)

## Any other provided annotations
tbl2 <- tbl |>
    dplyr::filter(RunID != "Shih")
if (nrow(tbl2) > 0 && (length(tbl2$SampleName) == length(unique(tbl2$SampleName)))) {
    for (cn in colnames(tbl2)) {
        if (!all(is.na(tbl2[[cn]])) && !(cn %in% colnames(colData(dat)))) {
            if (cn == "Day") {
                colData(dat)[[cn]] <- paste0(
                    "D", tbl2[[cn]][match(dat$sample, tbl2$SampleName)])
            } else if (cn == "RunID") {
                colData(dat)[[cn]] <- sub("_[0-9]+$", "", 
                                          tbl2[[cn]][match(dat$sample, 
                                                           tbl2$SampleName)])
            } else {
                colData(dat)[[cn]] <- tbl2[[cn]][match(dat$sample, tbl2$SampleName)]
            }
        }
    }
}

## Harmonize annotation values
dat$Stage <- gsub(" stage", "", dat$Stage)

## Add information for the Shih data set
shih_annot <- read.delim(shih_cellannot) |>
    dplyr::mutate(cell = paste0("Shih_", run, "_", gsub("-[0-9]$", "", barcode)))
if (any(dat$cell %in% shih_annot$cell)) {
    for (cn in c("clusterID", "pheno", "Phase", "subjectID", "NKsubclusterID",
                 "StromalSubclusterID")) {
        colData(dat)[[paste0("Shih_", cn)]] <- 
            shih_annot[[cn]][match(dat$cell, shih_annot$cell)]
    }
}
dat$RunID[grep("Shih", dat$sample)] <- "Shih"
dat$Stage[grep("Shih", dat$sample)] <- "menstrual"

dat$StageFactor <- factor(dat$Stage, 
                          levels = c("menstrual", "proliferative",
                                     "early-secretory", "early-mid-secretory",
                                     "mid-secretory", "late-secretory"))
dat$StageCoarse <- sub("early-|early-mid-|mid-|late-", "", dat$Stage)

## Create a 'donorCol' column
dat$donorCol <- NA_character_
if ("DonorID" %in% colnames(colData(dat))) {
    idx <- which(!is.na(dat[["DonorID"]]))
    dat$donorCol[idx] <- dat[["DonorID"]][idx]
}
if ("Shih_subjectID" %in% colnames(colData(dat))) {
    idx <- which(!is.na(dat[["Shih_subjectID"]]))
    dat$donorCol[idx] <- dat[["Shih_subjectID"]][idx]
}

## Create a 'sampleCol' column (either sample or Shih_subjectID)
dat$sampleCol <- NA_character_
if ("sample" %in% colnames(colData(dat))) {
    idx <- which(!is.na(dat[["sample"]]))
    dat$sampleCol[idx] <- dat[["sample"]][idx]
}
if ("Shih_subjectID" %in% colnames(colData(dat))) {
    idx <- which(!is.na(dat[["Shih_subjectID"]]))
    dat$sampleCol[idx] <- dat[["Shih_subjectID"]][idx]
}

## Generate columns to match to HECA
dat$matchDataset <- ifelse(dat$RunID %in% c("Wang", "Shih", "Lai", "Huang"), 
                           dat$RunID, "GarciaAlonso")
dat$matchCell <- sub("^Huang_|^Lai_|^Shih_", "", dat$cell)
dat$matchId <- paste0(dat$matchCell, "-", dat$matchDataset)

## Add annotations from the reproductive cell atlas (2021)
rca <- readRDS(rca)
table(dat$matchDataset, colnames(dat) %in% colnames(rca), useNA = "ifany")
dat$RCA_CellCyclePhase <- rca$`CellCycle Phase`[match(colnames(dat), colnames(rca))]
dat$RCA_Celltype <- rca$`Cell type`[match(colnames(dat), colnames(rca))]
dat$RCA_percentmito <- rca$percent_mito[match(colnames(dat), colnames(rca))]
dat$RCA_ngenes <- rca$n_genes[match(colnames(dat), colnames(rca))]
dat$RCA_BiopsyType <- rca$BiopsyType[match(colnames(dat), colnames(rca))]
dat$RCA_Location <- rca$Location[match(colnames(dat), colnames(rca))]
dat$RCA_BroadCelltype <- rca$`Broad cell type`[match(colnames(dat), colnames(rca))]

## Add annotations from HECA (2023)
heca <- readRDS(heca)
table(dat$matchDataset, dat$matchId %in% colnames(heca), useNA = "ifany")
dat$HECA_CellCyclePhase <- as.character(heca$phase[match(dat$matchId, colnames(heca))])
dat$HECA_Celltype <- as.character(heca$celltype[match(dat$matchId, colnames(heca))])
dat$HECA_ngenes <- heca$n_genes[match(dat$matchId, colnames(heca))]
dat$HECA_percentmito <- heca$percent_mito[match(dat$matchId, colnames(heca))]
dat$HECA_ScrubletScore <- heca$scrublet_score[match(dat$matchId, colnames(heca))]
dat$HECA_Stage <- as.character(heca$Stage[match(dat$matchId, colnames(heca))])
dat$HECA_Lineage <- as.character(heca$lineage[match(dat$matchId, colnames(heca))])
dat$HECA_Label <- as.character(heca$label_long[match(dat$matchId, colnames(heca))])

## Filter to only keep cells that have a value for the batch variable
(dat <- dat[, !is.na(dat[[batchvar]])])

## Sample annotation summary
DT::datatable(as.data.frame(colData(dat)) |>
                  dplyr::select(sample, SampleType, DonorID, SampleName, RunID,
                                ReadLength1, ReadLength2, Tissue, Stage,
                                Digestion, Protocol, Source, Sex, 
                                TimeInMenstrualCycle, DiseaseState, 
                                Fecundity, Phenotype, DevStage,
                                Shih_pheno, Shih_subjectID, 
                                StageFactor, StageCoarse, donorCol, sampleCol,
                                matchDataset) |>
                  dplyr::distinct(), 
              rownames = FALSE,
              options = list(scrollX = TRUE, pageLength = 50))
```

# Extract only epithelial cells for further analysis

We perform a basic pre-analysis with the aim to cluster the cells and extract
the ones that are likely of epithelial origin, for further processing. 

```{r select-epithelial}
set.seed(123)

## Normalize
dat0 <- scuttle::computeLibraryFactors(dat)
dat0 <- batchelor::multiBatchNorm(dat0, batch = dat0[[batchvar]], min.mean = 0.1)

## Find HVGs
scran_geneVar <- scran::modelGeneVar(dat0, block = dat0[[datasetvar]])
hvg.scran.genevar <- scran::getTopHVGs(scran_geneVar, n = 3000)

## Batch correction and dimension reduction
mnn.out <- fastMNN(dat0, batch = dat0[[batchvar]], d = 50, k = 20, 
                   subset.row = hvg.scran.genevar)
stopifnot(all(colnames(mnn.out) == colnames(dat0)))
reducedDim(dat0, "fastMNNcorr") <- reducedDim(mnn.out, "corrected")

dat0 <- scater::runUMAP(dat0, dimred = "fastMNNcorr", 
                        name = "UMAP_fastMNNcorr", n_neighbors = 50)
plotReducedDim(dat0, dimred = "UMAP_fastMNNcorr", colour_by = "HECA_Lineage", 
               point_size = 0.5) + 
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

## Cluster the UMAP and select the epithelial cluster
grph <- bluster::makeSNNGraph(reducedDim(dat0, "UMAP_fastMNNcorr"), 
                              k = 20, type = "rank", 
                              BPPARAM = MulticoreParam(workers = 8L))
clst <- igraph::cluster_louvain(grph, resolution = 0.001)
dat0$louvain <- as.character(clst$membership)
plotReducedDim(dat0, dimred = "UMAP_fastMNNcorr", colour_by = "louvain", 
               point_size = 0.5) + 
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

tbl <- table(dat0$louvain, dat0$HECA_Lineage)
tbl <- sweep(tbl, MARGIN = 1, STATS = rowSums(tbl), FUN = "/")
tbl

## epithelial
(keepclust_epithelial <- rownames(tbl)[tbl[, "Epithelial"] > 0.9])
dat0$keep_epithelial <- dat0$louvain %in% keepclust_epithelial
table(dat0$keep_epithelial)
scater::plotReducedDim(dat0, dimred = "UMAP_fastMNNcorr", 
                       colour_by = "keep_epithelial", point_size = 0.5) + 
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

## Subset the original data set to the epithelial cells, and delete the 
## temporary dataset used above
dat0 <- dat0[, which(dat0$keep_epithelial)]
dim(dat0)
stopifnot(colnames(dat0) %in% colnames(dat))
dat <- dat[, colnames(dat0)]
rm(dat0)
dat
```

# Check total count for each gene type and removed unwanted gene types

```{r filter-genes}
#| fig.width: 10
#| fig.height: 10

keepGeneTypes <- c("protein_coding", "IG_V_gene", "IG_C_gene", "IG_J_gene",
                   "TR_C_gene", "TR_J_gene", "TR_V_gene", "TR_D_gene",
                   "IG_D_gene", "Mt_tRNA", "Mt_rRNA")

pd <- data.frame(gene = rownames(dat), 
                 gene_type_ribo = rowData(dat)$gene_type_ribo, 
                 total_count = rowSums(counts(dat))) |>
    group_by(gene_type_ribo) |>
    summarize(nbr_genes = length(gene), 
              total_count = sum(total_count)) |>
    mutate(keep = gene_type_ribo %in% keepGeneTypes) |>
    arrange(desc(nbr_genes))
pd$gene_type_ribo <- factor(pd$gene_type_ribo, levels = unique(pd$gene_type_ribo))

p1 <- ggplot(pd, aes(x = gene_type_ribo, y = nbr_genes, fill = keep)) +
    geom_col() +
    labs(x = "", y = "Number of genes", fill = "keep") +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
p2 <- ggplot(pd, aes(x = gene_type_ribo, y = total_count, fill = keep)) +
    geom_col() +
    labs(x = "", y = "Total UMI count", fill = "keep") +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot_grid(p1, p2, ncol = 1)

dat <- dat[rowData(dat)$gene_type_ribo %in% keepGeneTypes, ]
dim(dat)
```

# Add QC metrics

```{r qc-scuttle}
(MT <- grep("^MT-", rownames(dat), value = TRUE))
dat <- scuttle::addPerCellQC(dat, assay.type = "counts", subsets = list(MT = MT))
dat <- scuttle::addPerFeatureQC(dat, assay.type = "counts")
dat$intronic_fraction <- 
    colSums(assay(dat, "unspliced")) / colSums(assay(dat, "total"))
rowData(dat)$intronic_fraction <- 
    rowSums(assay(dat, "unspliced")) / rowSums(assay(dat, "total"))
```

# QC plots

```{r qc-plots}
#| fig.width: 8
#| fig.height: 6

total_count_threshold <- 1000
detected_threshold <- 500
mt_percent_threshold <- 25

dat$keep <- (dat$total > total_count_threshold) & 
    (dat$detected > detected_threshold) & 
    (dat$subsets_MT_percent < mt_percent_threshold)
table(dat$keep)

cdplot <- as.data.frame(colData(dat))

sid <- "sample"
stid <- "StageFactor"

ggplot(cdplot, aes(x = .data[[sid]], y = sum, fill = .data[[stid]])) +
    geom_hline(yintercept = total_count_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Total UMI count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
ggplot(cdplot, aes(x = .data[[sid]], y = sum, fill = .data[[stid]])) +
    geom_hline(yintercept = total_count_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    scale_y_log10() + 
    labs(y = "Total UMI count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = .data[[sid]], y = detected, fill = .data[[stid]])) +
    geom_hline(yintercept = detected_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Number of detected genes per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = .data[[sid]], y = subsets_MT_percent, fill = .data[[stid]])) +
    geom_hline(yintercept = mt_percent_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Percentage of MT count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = detected, y = subsets_MT_percent, color = keep)) + 
    geom_point(size = 0.5, alpha = 0.5) + theme_minimal() + 
    labs(x = "Number of detected genes per cell", 
         y = "Percentage of MT count per cell")
ggplot(cdplot, aes(x = detected, y = intronic_fraction, color = keep)) + 
    geom_point(size = 0.5, alpha = 0.5) + theme_minimal() + 
    labs(x = "Number of detected genes per cell",
         y = "Intronic fraction per cell")
```

# Filter and remove samples with too few cells

```{r filter}
#| fig.width: 8
#| fig.height: 6

(tbl <- table(sample = dat$sample, keep = dat$keep))
ggplot(as.data.frame(tbl), aes(x = sample, y = Freq, fill = keep)) + 
    geom_col(position = "stack") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(x = "", y = "Number of cells")
ggplot(as.data.frame(tbl), aes(x = sample, y = Freq, fill = keep)) + 
    geom_col(position = "fill") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(x = "", y = "Fraction of cells")

dat <- dat[, which(dat$keep)]
dim(dat)
```

```{r cells-per-sample}
(tbl <- sort(table(dat$sampleCol)))
keepsamples <- names(tbl)[tbl > 30]
dat <- dat[, dat$sampleCol %in% keepsamples]
dim(dat)
```

```{r draw-heatmap}
#| fig.width: 8
#| fig.height: 7

tbl <- table(dat[[sid]], dat[[stid]])
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells per\nsample",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))
```

# Recalculate QC metrics for genes based on remaining cells

```{r qc-scuttle2}
dat <- scuttle::addPerFeatureQC(dat, assay.type = "counts")
rowData(dat)$intronic_fraction <- 
    rowSums(assay(dat, "unspliced")) / rowSums(assay(dat, "total"))
```

# Most highly expressed genes

```{r highly-expressed}
scater::plotHighestExprs(dat, n = 25, colour_cells_by = stid, 
                         exprs_values = "counts")
```

# Calculate size factors and logcounts

```{r size-factors}
size_factor_threshold <- 0.1
set.seed(123)
clusters <- scran::quickCluster(
    dat, method = "igraph", min.mean = 0.1,
    use.ranks = FALSE, BSPARAM = BiocSingular::ExactParam(),
    block = dat[[datasetvar]], 
    min.size = min(table(dat[[datasetvar]]))
)
dat <- scran::computeSumFactors(dat, min.mean = 0.1, cluster = clusters)
summary(sizeFactors(dat))
ggplot(data.frame(libSize = dat$sum,
                  sizeFactor = sizeFactors(dat),
                  group = dat[[blockvar]],
                  stringsAsFactors = FALSE),
       aes(x = libSize, y = sizeFactor, color = group)) + 
    geom_point(alpha = 0.5) + theme_bw() + 
    geom_hline(yintercept = size_factor_threshold, linetype = "dashed") + 
    xlab("Total UMI count") + ylab("Size factor")
dat <- dat[, sizeFactors(dat) > size_factor_threshold]
dim(dat)

dat <- batchelor::multiBatchNorm(dat, batch = dat[[batchvar]], min.mean = 0.1)
ggplot(data.frame(libSize = dat$sum,
                  sizeFactor = sizeFactors(dat),
                  group = dat[[blockvar]],
                  stringsAsFactors = FALSE),
       aes(x = libSize, y = sizeFactor, color = group)) + 
    geom_point(alpha = 0.5) + theme_bw() + 
    geom_hline(yintercept = size_factor_threshold, linetype = "dashed") + 
    xlab("Total UMI count") + ylab("Size factor")
```

# Cell cycle assignment

```{r cell-cycle}
#| message: false
#| warning: false

## tricycle
set.seed(123)
dat <- project_cycle_space(dat, species = "human", gname.type = "SYMBOL")
dat <- estimate_cycle_position(dat)
dat <- estimate_Schwabe_stage(dat, gname.type = "SYMBOL", species = "human", 
                              batch.v = batchvar, tolerance = 0)
plotReducedDim(dat, "tricycleEmbedding", point_size = 0.5, 
               colour_by = "CCStage")
tbl <- table(dat$StageFactor, dat$CCStage)
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))

## Seurat (using genes from HECA)
s_genes <- read.delim(s_genes)[, 1]
g2m_genes <- read.delim(g2m_genes)[, 1]
seurat <- CreateSeuratObject(counts = assay(dat, "counts"))
seurat <- NormalizeData(seurat, verbose = FALSE)
seurat <- FindVariableFeatures(seurat, selection.method = "vst", verbose = FALSE)
seurat <- ScaleData(seurat, features = rownames(seurat), verbose = FALSE)
seurat <- CellCycleScoring(seurat, s.features = s_genes, g2m.features = g2m_genes, 
                           set.ident = TRUE)
stopifnot(all(rownames(seurat@meta.data) == colnames(dat)))
dat$SeuratCC <- seurat@meta.data$Phase
tbl <- table(dat$StageFactor, dat$SeuratCC)
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))
```

# Find and mark doublets

```{r doublets}
#| warning: false

set.seed(123)
dat <- scDblFinder(dat, samples = blockvar, verbose = FALSE, 
                   BPPARAM = MulticoreParam(workers = 8L))
table(dat$scDblFinder.class, dat[[blockvar]])
```

# Sketching

We find highly variable genes while blocking on the batch, to focus on the 
biological variance.

```{r sketching}
set.seed(123)
sketchcells.bybatch <- unlist(lapply(unique(colData(dat)[[batchvar]]), function(b) {
    subd <- dat[, which(colData(dat)[[batchvar]] == b)]
    geneVar <- scran::modelGeneVar(subd)
    subhvgs <- scran::getTopHVGs(geneVar, n = 2000)
    subd <- scater::runPCA(subd, exprs_values = "logcounts", ncomponents = 30, 
                           name = "PCAsub", 
                           subset_row = subhvgs, scale = FALSE)
    idxsub <- geosketch(reducedDim(subd, "PCAsub"), 
                        N = min(ncol(subd), max(0.1 * ncol(subd), 150)), 
                        seed = 123)
    colnames(subd)[idxsub]
}))
dat$sketched.bybatch <- ifelse(colnames(dat) %in% sketchcells.bybatch, TRUE, FALSE)
table(batch = colData(dat)[[batchvar]], sketched.bybatch = dat$sketched.bybatch)
```

# Find an alternative reduced dimension representation, after sketching

```{r dimred-sketch}
set.seed(123)
## First find a new set of HVGs
dattmp <- dat[, sketchcells.bybatch]
stopifnot(all(rownames(dattmp) == rownames(dat)))
rowData(dat)$scran_geneVar_geosketch <- 
    scran::modelGeneVar(dattmp, block = dattmp[[datasetvar]])
hvgs <- scran::getTopHVGs(rowData(dat)$scran_geneVar_geosketch, n = 3000)
metadata(dat)$hvg.scran.genevar.geosketch <- hvgs

head(
    as.data.frame(
        rowData(dat)$scran_geneVar_geosketch[
            order(rowData(dat)$scran_geneVar_geosketch$bio, 
                  decreasing = TRUE), 
            c("mean", "total", "tech", "bio", "p.value", "FDR")]
    ), 
    25
)

## Next run multibatch PCA on the sketched cells
set.seed(123)
dattmp <- dat[, sketchcells.bybatch]
mbpca <- multiBatchPCA(dattmp, batch = dattmp[[batchvar]], d = 30, 
                       subset.row = metadata(dat)$hvg.scran.genevar.geosketch, 
                       weights = FALSE, preserve.single = TRUE)
projection <- mbpca[[1]]
rotation <- metadata(mbpca)$rotation
centers <- metadata(mbpca)$centers
## Sanity check
tmp <- t(sweep(assay(dattmp[names(centers), ], "logcounts"), MARGIN = 1, 
               STATS = centers, FUN = "-")) %*% rotation
max(abs(tmp - projection))
## Project all cells
pca <- t(sweep(assay(dat[names(centers), ], "logcounts"), MARGIN = 1, 
               STATS = centers, FUN = "-")) %*% rotation

## Perform batch correction in the PCA space
stopifnot(all(rownames(pca) == colnames(dat)))
reducedDim(dat, "multibatchPCA") <- as.matrix(pca)
mnn <- reducedMNN(as.matrix(pca), batch = dat[[batchvar]], k = 20, 
                  BPPARAM = MulticoreParam(workers = 8L))
stopifnot(all(rownames(mnn) == colnames(dat)))
reducedDim(dat, "fastMNNcorr_sketch") <- mnn$corrected
```

# Dimension reduction

Apply tSNE and UMAP to sketched cells and project the others.

```{r dimred}
## TSNE
set.seed(123)
tsne <- snifter::fitsne(
    reducedDim(dat[, sketchcells.bybatch], "fastMNNcorr_sketch"), 
    n_jobs = 8L, random_state = 123L,
    perplexity = 30, n_components = 2L, 
    pca = FALSE)
proj <- snifter::project(
    x = tsne, 
    new = reducedDim(dat, "fastMNNcorr_sketch"), 
    old = reducedDim(dat[, sketchcells.bybatch], "fastMNNcorr_sketch"))
rownames(proj) <- colnames(dat)
reducedDim(dat, "TSNE_fastMNNcorr_sketch") <- proj

## UMAP
set.seed(123)
umap <- uwot::umap(
    X = reducedDim(dat[, sketchcells.bybatch], "fastMNNcorr_sketch"), 
    n_components = 2, n_neighbors = 50,
    pca = NULL, scale = FALSE, 
    ret_model = TRUE, n_threads = 8L)
proj <- uwot::umap_transform(
    X = reducedDim(dat, "fastMNNcorr_sketch"), 
    model = umap, 
    n_threads = 8L)
rownames(proj) <- colnames(dat)
reducedDim(dat, "UMAP_fastMNNcorr_sketch") <- proj
```

# Shuffle cells (to avoid plotting in layers by sample later)

```{r shuffle-cells}
set.seed(123)
dat <- dat[, sample(seq_len(ncol(dat)), ncol(dat), replace = FALSE)]
```

# Clustering

```{r cluster}
## Helper function
## method should be "louvain" or "leiden"
clusterfcn <- function(sce, dimred = "fastMNNcorr", method = "leiden",
                       resolutions = c(0.05, 0.1, 0.2, 0.4, 0.6, 1.0), 
                       sketched_cells = NULL) {
    if (is.null(sketched_cells)) {
        for (r in resolutions) {
            set.seed(123)
            mat <- reducedDim(sce, dimred)
            clst <- bluster::clusterRows(
                mat, 
                BLUSPARAM = bluster::SNNGraphParam(
                    k = 10, type = "rank", 
                    num.threads = 1L, 
                    cluster.fun = method, 
                    cluster.args = list(resolution = r)))
            colData(sce)[[paste0(method, "_res", r)]] <- as.character(clst)
        }
    } else {
        train <- reducedDim(sce[, sketched_cells], dimred)
        test <- reducedDim(sce[, setdiff(colnames(sce), sketched_cells)], dimred)
        for (r in resolutions) {
            set.seed(123)
            df <- data.frame(cell = colnames(sce), cluster = NA_character_)
            clst <- as.character(bluster::clusterRows(
                train,  
                BLUSPARAM = bluster::SNNGraphParam(
                    k = 10, type = "rank", 
                    num.threads = 1L, 
                    cluster.fun = method, 
                    cluster.args = list(resolution = r))))
            preds <- class::knn(train = train, test = test, 
                                cl = clst, k = 1) |> as.character()
            df$cluster[match(rownames(train), df$cell)] <- clst
            df$cluster[match(rownames(test), df$cell)] <- preds
            stopifnot(colnames(sce) == df$cell)
            colData(sce)[[paste0(method, "_res", r)]] <- df$cluster
        }
    }
    
    sce
}

dat <- clusterfcn(dat, dimred = "fastMNNcorr_sketch", method = "leiden",
                  resolutions = 0.4, 
                  sketched_cells = sketchcells.bybatch)
```

```{r replace-clusters}
#| echo: false

## For some reason, sometimes the clustering above does not respect the 
## random seed. For reproducibility, we therefore saved the cluster annotations
## to a file. We will read it here, and replace the cluster labels if needed
cluster_labels <- read.delim("leiden_clustering_0.4.tsv",  header = TRUE)
if (all(dat$cell == cluster_labels$cell)) {
    if (any(dat$leiden_res0.4 != cluster_labels$leiden_res0.4)) {
        print("Replacing cluster labels")
        dat$leiden_res0.4 <- as.character(cluster_labels$leiden_res0.4)
    }
} else {
    print("Cells are not in the same order as in the cluster annotation")
}
```

# Annotate clusters

```{r annotate-clusters}
clust_annot <- stack(
    list("pre-ciliated" = 2, 
         "ciliated" = c(1, 3, 8, 30), 
         "secretory late" = 41, 
         "secretory mid" = c(22, 25, 35, 34, 40, 38, 21, 18, 19, 23), 
         "secretory early" = c(26, 31, 36, 37), 
         "luminal late" = c(45, 28, 32, 46, 20, 33), 
         "transcriptionally active" = c(15, 10, 39), 
         "luminal early" = 14, 
         "MUC5B+" = 42, 
         "KRT5+" = 43, 
         "remodeling" = c(16, 24, 13), 
         "cycling" = c(4, 29, 6, 11, 12, 9), 
         "hormone responsive early" = c(5, 7, 47), 
         "hormone responsive late" = c(17, 44, 27))
)
dat$leiden_res0.4_annot <- clust_annot$ind[match(dat$leiden_res0.4, 
                                                 as.character(clust_annot$values))]
table(dat$leiden_res0.4_annot)

leiden_res0.4_annot_cols <- c(
    "pre-ciliated" = "lightskyblue", 
    "ciliated" = "royalblue", 
    "secretory early" = "#C8AEE2", 
    "secretory mid" = "#7D4DBA", 
    "secretory late" = "#3A1260", 
    "luminal early" = "gold1", 
    "luminal late" = "darkorange", 
    "transcriptionally active" = "turquoise4", 
    "MUC5B+" = "salmon", 
    "KRT5+" = "firebrick", 
    "remodeling" = "#4CAF50", 
    "cycling" = "mediumaquamarine", 
    "hormone responsive early" = "hotpink2", 
    "hormone responsive late" = "hotpink4" 
) 
```

# Dot plots of clusters and marker genes of interest

```{r markers1}
#| fig.width: 8
#| fig.height: 10
#| warning: false

col = circlize::colorRamp2(
    breaks = seq(0, 7.5, length.out = 64),
    colors = colorRampPalette(c("grey90", "purple"))(64))

markers <- intersect(
    c("MKI67", "MMP7", "STMN1", "PKM", "ALDH1A1", "VIM", "LGR5", "LEFTY1",
      "IL6", "ENPP3", "WNT7A", "IL32", "SAA1", "SAA2", "CLU", "MUC5B", "LCN2", 
      "LY6E", "S100A6", "PLAUR", "PLAU", "SFN", "TNFRSF12A", "IL23A", "KRT17",
      "KRT5", "ANXA1", "CXCL8", "SCGB2A1", "SBGB2A2", "SCGB1D2", "IL6ST", 
      "PAEP", "FGF7", "PGR", "ESR1", "CCNO", "FOXJ1", "PIFO", "RSPH1", "DKK1"),
    rownames(dat))
plotDots(dat, features = markers,
         group = "leiden_res0.4_annot", 
         swap_rownames = "gene_name", zlim = c(0, 7.5),
         colour = col(seq(0, 7.5, length.out = 64))) + 
    scale_size(limits=c(0.2, 1)) + 
    labs(x = "") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Marker genes from HECA paper (https://www.nature.com/articles/s41588-024-01873-w, Fig 2a).

```{r markers2}
#| fig.width: 8
#| fig.height: 12
#| warning: false

markers <- intersect(
    c("PAX8", "EPCAM", "FOXA2", "MMP7", "SOX9", "AXIN2", "TRH", "ALDH1A1",
      "CDH2", "KLK11", "SLC7A11", "DKK1", "PHLDA1", "KMO", "IHH", "EMID1", 
      "HPRT1", "SUFU", "OPRK1", "CBR3", "S100P", "SCGB2A2", "ABCG1", "PAEP", 
      "DPP4", "GPX3", "PTPRR", "FGF7", "FXYD2", "IL32", "WNT7A", "LGR5", 
      "MMP7", "TNF", "VTCN1", "CLDN22", "SULT1E1", "LEFTY1", "PTGS1", "IL6",
      "CCNO", "CDC20B", "MUC12", "PIFO", "FOXJ1", "TP73", "BPIFB1", "TFF3", 
      "MUC5B", "SAA1", "TP63", "SNCG", "KRT5", "PGR", "AR", "ESR1"), 
    rownames(dat))
plotDots(dat, features = markers,
         group = "leiden_res0.4_annot", 
         swap_rownames = "gene_name", zlim = c(0, 7.5),
         colour = col(seq(0, 7.5, length.out = 64))) + 
    scale_size(limits=c(0.2, 1)) + 
    labs(x = "") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


# Plot dimension reductions

```{r plot-umap}
#| fig.width: 8
#| fig.height: 6

stage_colors <- c( 
    "early-secretory" = "#F4EBFA",
    "early-mid-secretory" = "#C8AEE2",
    "proliferative" = "hotpink2",
    "mid-secretory" = "#7D4DBA",
    "late-secretory" = "#3A1260",
    "menstrual" = "#A30B2E"
)

seurat_colors <- c(
    "S" = "gold",
    "G1" = "grey80",
    "G2M" = "brown"
)

plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
               other_fields = "HECA_Celltype", point_size = 0.5) + 
    geom_point(aes(color = HECA_Celltype), size = 0.5) + 
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
    theme_void()

for (vrb in c("RCA_Celltype", "RCA_Location", "RCA_BiopsyType", 
              "matchDataset", "StageCoarse", "donorCol", 
              "scDblFinder.class", "leiden_res0.4")) {
    print(plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
                         colour_by = vrb, point_size = 0.5) + 
              guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
              theme_void())
}

print(plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
                     colour_by = "leiden_res0.4_annot", point_size = 0.5) + 
          scale_color_manual(values = leiden_res0.4_annot_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())
print(plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
                     colour_by = "Stage", point_size = 0.5) + 
          scale_color_manual(values = stage_colors, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())
print(plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
                     colour_by = "SeuratCC", point_size = 0.5) + 
          scale_color_manual(values = seurat_colors, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())

for (vrb in c("subsets_MT_percent", "detected", "intronic_fraction")) {
    print(plotReducedDim(dat, dimred = "UMAP_fastMNNcorr_sketch", 
                         colour_by = vrb, point_size = 0.5) + 
              theme_void())
}

```

# Plot cluster abundance over the cycle

```{r plot-over-time}
#| warning: false
#| message: false
#| fig.height: 8
#| fig.width: 9.5

col <- "leiden_res0.4_annot"

df <- data.frame(colData(dat)[, c("StageFactor", col)]) |>
    group_by(StageFactor, .data[[col]]) |>
    tally() |>
    group_by(StageFactor) |>
    mutate(perc = n / sum(n) * 100) |>
    group_by(.data[[col]]) |>
    mutate(scaledperc = perc / sum(perc)) |>
    ungroup() |>
    dplyr::filter(!is.na(StageFactor))
ggplot(df, aes(x = StageFactor, y = perc, color = .data[[col]])) + 
    geom_point() + 
    geom_line(aes(x = as.numeric(StageFactor)), linewidth = 1) + 
    theme_cowplot() + 
    labs(x = "", y = "Relative abundance (%)",
         title = col) + 
    theme(legend.position = "bottom") + 
    scale_color_manual(values = leiden_res0.4_annot_cols, name = "")

## One panel per cluster
ggplot(df, aes(x = StageFactor, y = perc, color = .data[[col]])) + 
    geom_point() + 
    geom_line(aes(x = as.numeric(StageFactor)), linewidth = 1) + 
    theme_cowplot() + 
    labs(x = "", y = "Relative abundance (%)",
         title = col) + 
    facet_wrap(~ .data[[col]], scales = "free_y") + 
    scale_color_manual(values = leiden_res0.4_annot_cols, name = "") + 
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.text = element_text(size = 10))
```

# Find marker genes

First, we compare all clusters in a pairwise manner. The comparison is only 
performed for genes with a non-zero count in at least 
0.05% of the cells (corresponding to `r ceiling(0.05/100*ncol(dat))` of the 
`r ncol(dat)` cells in the data set). These gene IDs are also written to a 
file, for potential use as the 'universe' in over-representation analyses. 

```{r}
#| warning: false

frac_detected <- rowMeans(assay(dat, "counts") > 0)
genes_to_test <- rownames(dat)[frac_detected >= (0.05 / 100)]
write.table(genes_to_test, file = "epithelial_cell_atlas_gene_universe.txt", 
            row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")

clc <- "leiden_res0.4_annot"
ptts <- pairwiseTTests(dat, groups = dat[[clc]], 
                       block = dat[[blockvar]], direction = "up", 
                       subset.row = genes_to_test)
markers <- combineMarkers(ptts$statistics, ptts$pairs, pval.type = "all")

## Write results to files
ttestdir <- paste0("pairwise_ttests_", clc)
if (!file.exists(ttestdir)) {
    dir.create(ttestdir, recursive = TRUE)
}
for (i in seq_len(nrow(ptts$pairs))) {
    df <- cbind(gene = rownames(ptts$statistics[[i]]), 
                ptts$statistics[[i]])
    df <- df[order(df$p.value, -df$logFC), ]
    filename <- paste0(clc, "_cluster_", ptts$pairs$first[i],
                       "_vs_", ptts$pairs$second[i], ".txt")
    write.table(df, file = file.path(ttestdir, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}

markerdir <- paste0("marker_genes_all_", clc)
if (!file.exists(markerdir)) {
    dir.create(markerdir, recursive = TRUE)
}
for (i in seq_along(markers)) {
    df <- cbind(gene = rownames(markers[[i]]), markers[[i]])
    filename <- paste0(clc, "_cluster_", names(markers)[i], "_markers.txt")
    write.table(df, file = file.path(markerdir, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}
```

# Save object

```{r save-object}
saveRDS(dat, file = "invivo_epithelial_cell_atlas.rds")
```

# Session info

```{r session-info}
sessionInfo()
```


