---
title: "Organoid scRNAseq timecourse"
date: "`r Sys.Date()`"
format: 
    html:
        toc: true
        theme: united
embed-resources: true
editor_options: 
  chunk_output_type: console
---

# Load packages

```{r}
suppressPackageStartupMessages({
    library(fishpond)
    library(SummarizedExperiment)
    library(SingleCellExperiment)
    library(scuttle)
    library(scater)
    library(scran)
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(msigdbr)
    library(cowplot)
    library(ComplexHeatmap)
    library(circlize)
    library(scDblFinder)
    library(Seurat)
    library(sketchR)
    library(tricycle)
    library(org.Hs.eg.db)
    library(parallel)
    library(bluster)
    library(BiocParallel)
    library(igraph)
    library(DT)
    library(swissknife)
    library(SingleR)
})
```

# Define parameters and reference files

```{r}
## Sample annotation
sampleannot <- "scrnaseq_timecourse_samples.txt"

## Gene annotation (derived from GENCODE v38 gtf file)
t2g <- "reference/gencode.v38.annotation.tx2gene_full.rds"

## Cell cycle marker genes
## https://github.com/ventolab/HECA-Human-Endometrial-Cell-Atlas/tree/main/utils
s_genes <- "data/reproductivecellatlas/2023/S_genes_HECA.tsv"
g2m_genes <- "data/reproductivecellatlas/2023/G2M_genes_HECA.tsv"

## Tissue atlas
tissue_atlas_sce <- "../scRNAseq_invivo_endometrium_epithelial_cell_atlas/invivo_epithelial_cell_atlas.rds"
```

# Assemble gene annotation

```{r}
geneannot <- readRDS(t2g) |>
    dplyr::select(gene_id, seqnames, gene_type, gene_name) |>
    dplyr::rename(chromosome = seqnames) |>
    dplyr::distinct()

## Annotate ribosomal protein genes
## GO term "structural constituent of ribosome" (GO:0003735)
golist <- as.list(org.Hs.egGO2ALLEGS)
ribo_entrez <- unique(golist[["GO:0003735"]])
length(ribo_entrez)
tmp <- select(org.Hs.eg.db, keys = ribo_entrez, columns = "ENSEMBL", 
              keytype = "ENTREZID")
ribo_ensembl <- unique(na.omit(tmp[, "ENSEMBL"]))
length(ribo_ensembl)
summary(ribo_ensembl %in% sub("\\.[0-9]+$", "", geneannot$gene_id))
geneannot$is_ribosomal_protein <- 
    sub("\\.[0-9]+$", "", geneannot$gene_id) %in% ribo_ensembl
geneannot$gene_type_ribo <- ifelse(
    geneannot$gene_type == "protein_coding",
    ifelse(geneannot$is_ribosomal_protein, "protein_coding_ribosome",
           geneannot$gene_type),
    geneannot$gene_type)
```

# Read data

```{r}
#| message: false

## Get sample IDs
tbl <- read.delim(sampleannot)
DT::datatable(tbl, options = list(scrollX = TRUE, pageLength = 50))

## Read alevin-fry data and combine
dat <- do.call(cbind, lapply(seq_len(nrow(tbl)), function(i) {
    x <- fishpond::loadFry(
        fryDir = file.path("data", tbl$dir[i], tbl$sample[i]), 
        outputFormat = list(counts = c("S", "A"), spliced = c("S", "A"),
                            unspliced = c("U"), total = c("U", "S", "A")))
    stopifnot(all(!duplicated(colnames(x))))
    rowData(x)$gene_name <- 
        geneannot$gene_name[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$chromosome <- 
        geneannot$chromosome[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$gene_type <- 
        geneannot$gene_type[match(rowData(x)$gene_ids, geneannot$gene_id)]
    rowData(x)$gene_type_ribo <- 
        geneannot$gene_type_ribo[match(rowData(x)$gene_ids, geneannot$gene_id)]
    
    ## Keep only cells with at least 1,000 detected genes
    x <- x[, colSums(assay(x, "total") > 0) > 1000]
    
    ## Add sample ID to colnames
    x$sample <- tbl$sample[i]
    colnames(x) <- paste0(tbl$sample[i], "_", colnames(x))
    x
}))
dat

## Set rownames to gene names where possible
rownames(dat) <- scater::uniquifyFeatureNames(
    ID = rownames(dat), 
    names = rowData(dat)$gene_name
)
```

# Add sample annotations

```{r}
## Add 'cell' column
dat$cell <- colnames(dat)

## Any other annotations
if (nrow(tbl) > 0 && (length(tbl$sample) == length(unique(tbl$sample)))) {
    for (cn in colnames(tbl)) {
        if (!all(is.na(tbl[[cn]])) && !(cn %in% colnames(colData(dat)))) {
            colData(dat)[[cn]] <- tbl[[cn]][match(dat$sample, tbl$sample)]
        }
    }
}

## Create donorCol
dat$donorCol <- dat$donor

## Order timepoints
dat$timepoint <- c(d8 = "Horm Withdr 48h", d9_12h = "Post-breakdown 12h", 
                   d9_24h = "Post-breakdown 24h", d10 = "Post-breakdown 48h", 
                   d11 = "E2-diff 24h", d12 = "E2-diff 48h")[dat$timepoint]
dat$timepoint <- factor(dat$timepoint, 
                        levels = c("Horm Withdr 48h", "Post-breakdown 12h", 
                                   "Post-breakdown 24h", "Post-breakdown 48h", 
                                   "E2-diff 24h", "E2-diff 48h"))
table(dat$timepoint, useNA = "ifany")
```

# Check total count for each gene type and removed unwanted gene types

```{r}
#| fig.width: 10
#| fig.height: 10

keepGeneTypes <- c("protein_coding", "IG_V_gene", "IG_C_gene", "IG_J_gene",
                   "TR_C_gene", "TR_J_gene", "TR_V_gene", "TR_D_gene",
                   "IG_D_gene", "Mt_tRNA", "Mt_rRNA")

pd <- data.frame(gene = rownames(dat), 
                 gene_type_ribo = rowData(dat)$gene_type_ribo, 
                 total_count = rowSums(counts(dat))) |>
    group_by(gene_type_ribo) |>
    summarize(nbr_genes = length(gene), 
              total_count = sum(total_count)) |>
    mutate(keep = gene_type_ribo %in% keepGeneTypes) |>
    arrange(desc(nbr_genes))
pd$gene_type_ribo <- factor(pd$gene_type_ribo, levels = unique(pd$gene_type_ribo))

p1 <- ggplot(pd, aes(x = gene_type_ribo, y = nbr_genes, fill = keep)) +
    geom_col() +
    labs(x = "", y = "Number of genes", fill = "keep") +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
p2 <- ggplot(pd, aes(x = gene_type_ribo, y = total_count, fill = keep)) +
    geom_col() +
    labs(x = "", y = "Total UMI count", fill = "keep") +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot_grid(p1, p2, ncol = 1)

dat <- dat[rowData(dat)$gene_type_ribo %in% keepGeneTypes, ]
dim(dat)
```

# Add QC metrics

```{r qc-scuttle}
(MT <- grep("^MT-", rownames(dat), value = TRUE))
dat <- scuttle::addPerCellQC(dat, assay.type = "counts", subsets = list(MT = MT))
dat <- scuttle::addPerFeatureQC(dat, assay.type = "counts")
dat$intronic_fraction <- 
    colSums(assay(dat, "unspliced")) / colSums(assay(dat, "total"))
rowData(dat)$intronic_fraction <- 
    rowSums(assay(dat, "unspliced")) / rowSums(assay(dat, "total"))
```

# QC plots

```{r qc-plots}
#| fig.width: 8
#| fig.height: 6

total_count_threshold <- 1000
detected_threshold <- 500
mt_percent_threshold <- 25

dat$keep <- (dat$total > total_count_threshold) & 
    (dat$detected > detected_threshold) & 
    (dat$subsets_MT_percent < mt_percent_threshold)
table(dat$keep)

cdplot <- as.data.frame(colData(dat))

sid <- "sample"
stid <- "timepoint"

ggplot(cdplot, aes(x = .data[[sid]], y = sum, fill = .data[[stid]])) +
    geom_hline(yintercept = total_count_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Total UMI count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
ggplot(cdplot, aes(x = .data[[sid]], y = sum, fill = .data[[stid]])) +
    geom_hline(yintercept = total_count_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    scale_y_log10() + 
    labs(y = "Total UMI count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = .data[[sid]], y = detected, fill = .data[[stid]])) +
    geom_hline(yintercept = detected_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Number of detected genes per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = .data[[sid]], y = subsets_MT_percent, fill = .data[[stid]])) +
    geom_hline(yintercept = mt_percent_threshold) + 
    geom_violin(scale = "width") + theme_minimal() +
    labs(y = "Percentage of MT count per cell") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(cdplot, aes(x = detected, y = subsets_MT_percent, color = keep)) + 
    geom_point(size = 0.5, alpha = 0.5) + theme_minimal() + 
    labs(x = "Number of detected genes per cell",
         y = "Percentage of MT count per cell")
ggplot(cdplot, aes(x = detected, y = intronic_fraction, color = keep)) + 
    geom_point(size = 0.5, alpha = 0.5) + theme_minimal() + 
    labs(x = "Number of detected genes per cell",
         y = "Intronic fraction per cell")
```

# Filter and remove samples with too few cells

```{r}
#| fig.width: 8
#| fig.height: 6

(tbl <- table(sample = dat$sample, keep = dat$keep))
ggplot(as.data.frame(tbl), aes(x = sample, y = Freq, fill = keep)) + 
    geom_col(position = "stack") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(x = "", y = "Number of cells")
ggplot(as.data.frame(tbl), aes(x = sample, y = Freq, fill = keep)) + 
    geom_col(position = "fill") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(x = "", y = "Fraction of cells")

dat <- dat[, which(dat$keep)]
dim(dat)
```

```{r}
#| fig.width: 8
#| fig.height: 7

tbl <- table(dat[[sid]], dat[[stid]])
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells per\nsample",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))
```

# Recalculate QC metrics for genes based on remaining cells

```{r qc-scuttle-2}
dat <- scuttle::addPerFeatureQC(dat, assay.type = "counts")
rowData(dat)$intronic_fraction <- 
    rowSums(assay(dat, "unspliced")) / rowSums(assay(dat, "total"))
```

# Most highly expressed genes

```{r highly-expressed}
scater::plotHighestExprs(dat, n = 25, colour_cells_by = stid, 
                         exprs_values = "counts")
```

# Calculate size factors and logcounts

```{r size-factors}
size_factor_threshold <- 0.1
set.seed(123)
clusters <- scran::quickCluster(
    dat, method = "igraph", min.mean = 0.1,
    use.ranks = FALSE, BSPARAM = BiocSingular::ExactParam()
)
dat <- scran::computeSumFactors(dat, min.mean = 0.1, cluster = clusters)
summary(sizeFactors(dat))
ggplot(data.frame(libSize = dat$sum,
                  sizeFactor = sizeFactors(dat),
                  group = dat$sample,
                  stringsAsFactors = FALSE),
       aes(x = libSize, y = sizeFactor, color = group)) + 
    geom_point(alpha = 0.5) + theme_bw() + 
    geom_hline(yintercept = size_factor_threshold, linetype = "dashed") + 
    xlab("Total UMI count") + ylab("Size factor")
dat <- dat[, sizeFactors(dat) > size_factor_threshold]
dim(dat)

dat <- logNormCounts(dat)
```

# Cell cycle assignment

```{r}
#| message: false
#| warning: false

## tricycle
set.seed(123)
dat <- project_cycle_space(dat, species = "human", gname.type = "SYMBOL")
dat <- estimate_cycle_position(dat)
dat <- estimate_Schwabe_stage(dat, gname.type = "SYMBOL", species = "human", 
                              tolerance = 0)
plotReducedDim(dat, "tricycleEmbedding", point_size = 0.5, 
               colour_by = "CCStage")
tbl <- table(dat$timepoint, dat$CCStage)
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))

## Seurat (using genes from HECA)
s_genes <- read.delim(s_genes)[, 1]
g2m_genes <- read.delim(g2m_genes)[, 1]
seurat <- CreateSeuratObject(counts = assay(dat, "counts"))
seurat <- NormalizeData(seurat, verbose = FALSE)
seurat <- FindVariableFeatures(seurat, selection.method = "vst", verbose = FALSE)
seurat <- ScaleData(seurat, features = rownames(seurat), verbose = FALSE)
seurat <- CellCycleScoring(seurat, s.features = s_genes, g2m.features = g2m_genes, 
                           set.ident = TRUE)
stopifnot(all(rownames(seurat@meta.data) == colnames(dat)))
dat$SeuratCC <- seurat@meta.data$Phase
tbl <- table(dat$timepoint, dat$SeuratCC)
draw(Heatmap(tbl, col = circlize::colorRamp2(c(0, max(tbl)), c("white", "steelblue")),
             cluster_columns = FALSE, cluster_rows = FALSE, 
             row_names_gp = gpar(fontsize = 8),
             name = "Number of\ncells",
             rect_gp = gpar(col = "grey", lwd = 1),
             cell_fun = function(j, i, x, y, width, height, fill) {
                 grid.text(sprintf("%d", tbl[i, j]), x, y, gp = gpar(fontsize = 8))
             }))
```

# Find and mark doublets

```{r doublets}
#| warning: false

set.seed(123)
dat <- scDblFinder(dat, samples = "name", verbose = FALSE, 
                   BPPARAM = MulticoreParam(workers = 8L))
table(dat$scDblFinder.class, dat$name)
```

# Sketching

```{r hvg}
set.seed(123)
geneVar <- scran::modelGeneVar(dat)
subhvgs <- scran::getTopHVGs(geneVar, n = 2000)
subd <- scater::runPCA(dat, exprs_values = "logcounts", ncomponents = 30, 
                       name = "PCAsub", 
                       subset_row = subhvgs, scale = FALSE)
idxsub <- geosketch(reducedDim(subd, "PCAsub"), 
                    N = min(ncol(subd), max(0.1 * ncol(subd), 150)), 
                    seed = 123)
sketchcells <- colnames(subd)[idxsub]

dat$sketched <- ifelse(colnames(dat) %in% sketchcells, TRUE, FALSE)
table(sketched = dat$sketched)
```

# Find an alternative reduced dimension representation, after sketching

```{r}
set.seed(123)
## First find a new set of HVGs
dattmp <- dat[, sketchcells]
stopifnot(all(rownames(dattmp) == rownames(dat)))
rowData(dat)$scran_geneVar_geosketch <- scran::modelGeneVar(dattmp)
hvgs <- scran::getTopHVGs(rowData(dat)$scran_geneVar_geosketch, n = 3000)
metadata(dat)$hvg.scran.genevar.geosketch <- hvgs

head(
    as.data.frame(
        rowData(dat)$scran_geneVar_geosketch[
            order(rowData(dat)$scran_geneVar_geosketch$bio, 
                  decreasing = TRUE), 
            c("mean", "total", "tech", "bio", "p.value", "FDR")]
    ), 
    25
)

## Next run PCA on the sketched cells
set.seed(123)
dattmp <- dat[, sketchcells]
pca <- calculatePCA(dattmp, ncomponents = 30, 
                    subset_row = metadata(dat)$hvg.scran.genevar.geosketch)
projection <- pca
rotation <- attr(pca, "rotation")
centers <- rowMeans(assay(dattmp[metadata(dat)$hvg.scran.genevar.geosketch, ],
                          "logcounts"))
## Sanity check
stopifnot(rownames(rotation) == names(centers))
tmp <- t(sweep(assay(dattmp[names(centers), ], "logcounts"), MARGIN = 1, 
               STATS = centers, FUN = "-")) %*% rotation
max(abs(tmp - projection))
## Project all cells
pca <- t(sweep(assay(dat[names(centers), ], "logcounts"), MARGIN = 1, 
               STATS = centers, FUN = "-")) %*% rotation

stopifnot(all(rownames(pca) == colnames(dat)))
reducedDim(dat, "PCA_sketch") <- as.matrix(pca)
```

# Dimension reduction

Apply tSNE and UMAP to sketched cells and project the others.

```{r}
## TSNE
set.seed(123)
tsne <- snifter::fitsne(
    reducedDim(dat[, sketchcells], "PCA_sketch"), 
    n_jobs = 8L, random_state = 123L,
    perplexity = 30, n_components = 2L, 
    pca = FALSE)
proj <- snifter::project(
    x = tsne, 
    new = reducedDim(dat, "PCA_sketch"), 
    old = reducedDim(dat[, sketchcells], "PCA_sketch"))
rownames(proj) <- colnames(dat)
reducedDim(dat, "TSNE_sketch") <- proj

## UMAP
set.seed(123)
umap <- uwot::umap(
    X = reducedDim(dat[, sketchcells], "PCA_sketch"), 
    n_components = 2, n_neighbors = 50,
    pca = NULL, scale = FALSE, 
    ret_model = TRUE, n_threads = 8L)
proj <- uwot::umap_transform(
    X = reducedDim(dat, "PCA_sketch"), 
    model = umap, 
    n_threads = 8L)
rownames(proj) <- colnames(dat)
reducedDim(dat, "UMAP_sketch") <- proj
```

# Shuffle cells (to avoid plotting in layers by sample later)

```{r shuffle-cells}
set.seed(123)
dat <- dat[, sample(seq_len(ncol(dat)), ncol(dat), replace = FALSE)]
```

# Clustering

```{r}
## Helper function
## method should be "louvain" or "leiden"
clusterfcn <- function(sce, dimred = "fastMNNcorr", method = "leiden",
                       resolutions = c(0.05, 0.1, 0.2, 0.4, 0.6, 1.0), 
                       sketched_cells = NULL) {
    if (is.null(sketched_cells)) {
        for (r in resolutions) {
            set.seed(123)
            mat <- reducedDim(sce, dimred)
            clst <- bluster::clusterRows(
                mat, 
                BLUSPARAM = bluster::SNNGraphParam(
                    k = 10, type = "rank", 
                    num.threads = 1L, 
                    cluster.fun = method, 
                    cluster.args = list(resolution = r)))
            colData(sce)[[paste0(method, "_res", r)]] <- as.character(clst)
        }
    } else {
        train <- reducedDim(sce[, sketched_cells], dimred)
        test <- reducedDim(sce[, setdiff(colnames(sce), sketched_cells)], dimred)
        for (r in resolutions) {
            set.seed(123)
            df <- data.frame(cell = colnames(sce), cluster = NA_character_)
            clst <- as.character(bluster::clusterRows(
                train,  
                BLUSPARAM = bluster::SNNGraphParam(
                    k = 10, type = "rank", 
                    num.threads = 1L, 
                    cluster.fun = method, 
                    cluster.args = list(resolution = r))))
            preds <- class::knn(train = train, test = test, 
                                cl = clst, k = 1) |> as.character()
            df$cluster[match(rownames(train), df$cell)] <- clst
            df$cluster[match(rownames(test), df$cell)] <- preds
            stopifnot(colnames(sce) == df$cell)
            colData(sce)[[paste0(method, "_res", r)]] <- df$cluster
        }
    }
    
    sce
}

dat <- clusterfcn(dat, dimred = "PCA_sketch", method = "leiden",
                  resolutions = c(0.125), 
                  sketched_cells = sketchcells)
```

```{r}
#| echo: false

## For some reason, sometimes the clustering above does not respect the 
## random seed. For reproducibility, we therefore saved the cluster annotations
## to a file. We will read it here, and replace the cluster labels if needed
# write.table(colData(dat)[, c("cell", "leiden_res0.125")], 
#             file = "leiden_clustering.tsv", quote = FALSE, sep = "\t", 
#             row.names = FALSE, col.names = TRUE)
cluster_labels <- read.delim("leiden_clustering.tsv",  header = TRUE)
if (all(dat$cell == cluster_labels$cell)) {
    if (any(dat$leiden_res0.125 != cluster_labels$leiden_res0.125)) {
        print("Replacing cluster labels")
        dat$leiden_res0.125 <- as.character(cluster_labels$leiden_res0.125)
    }
} else {
    print("Cells are not in the same order as in the cluster annotation")
}
```

# Annotate clusters

```{r annotate-clusters}
clust_annot <- stack(
    list("pre-ciliated" = 7,
         "ciliated stress responsive" = 24,
         "ciliated wound responsive 12h" = c(23, 15, 27),
         "ciliated wound responsive 24h" = 6,
         "ciliated immunomodulatory" = c(25, 8), 
         "ciliated cycling" = 26,
         "ciliated" = c(1, 5, 3, 12, 16),
         "cycling" = 2,
         "stress responsive" = 17,
         "wound responsive 12h" = c(19, 20),
         "immunomodulatory" = 9,
         "wound responsive 24h" = c(10, 11, 14, 22),
         "OxPhos high" = 21,
         "secretory late" = 18,
         "secretory early" = 4,
         "hormone responsive" = 13)
)
dat$leiden_res0.125_annot <- 
    clust_annot$ind[match(dat$leiden_res0.125, 
                          as.character(clust_annot$values))]
table(dat$leiden_res0.125_annot)

leiden_res0.125_annot_cols <- c(
    "secretory late" = "firebrick",
    "secretory early" = "firebrick1",
    "hormone responsive" = "orchid3",
    "cycling" = "forestgreen",
    "stress responsive" = "grey20", 
    "wound responsive 12h" = "navajowhite",
    "wound responsive 24h" = "goldenrod1",
    "OxPhos high" = "coral4",
    "immunomodulatory" = "#FF6347", 
    "pre-ciliated" = "lightskyblue",
    "ciliated" = "royalblue",
    "ciliated stress responsive" = "navy",
    "ciliated wound responsive 12h" = "turquoise",
    "ciliated wound responsive 24h" = "turquoise4", 
    "ciliated immunomodulatory" = "tan1",  
    "ciliated cycling" = "olivedrab2"
) 

atlas_leiden_res0.4_annot_cols <- c(
    "pre-ciliated" = "lightskyblue", 
    "ciliated" = "royalblue", 
    "secretory early" = "#C8AEE2", 
    "secretory mid" = "#7D4DBA", 
    "secretory late" = "#3A1260", 
    "luminal early" = "gold1", 
    "luminal late" = "darkorange", 
    "transcriptionally active" = "turquoise4", 
    "MUC5B+" = "salmon", 
    "KRT5+" = "firebrick", 
    "remodeling" = "#4CAF50", 
    "cycling" = "mediumaquamarine", 
    "hormone responsive early" = "hotpink2", 
    "hormone responsive late" = "hotpink4" 
) 
```

# Dot plots of clusters and marker genes of interest

```{r markers1}
#| fig.width: 8
#| fig.height: 10
#| warning: false

col = circlize::colorRamp2(
    breaks = seq(0, 7.5, length.out = 64),
    colors = colorRampPalette(c("grey90", "purple"))(64))

markers <- intersect(
    c("MKI67", "MMP7", "STMN1", "PKM", "ALDH1A1", "VIM", "LGR5", "LEFTY1",
      "IL6", "ENPP3", "WNT7A", "IL32", "SAA1", "SAA2", "CLU", "MUC5B", "LCN2", 
      "LY6E", "S100A6", "PLAUR", "PLAU", "SFN", "TNFRSF12A", "IL23A", "KRT17",
      "KRT5", "ANXA1", "CXCL8", "SCGB2A1", "SBGB2A2", "SCGB1D2", "IL6ST", 
      "PAEP", "FGF7", "PGR", "ESR1", "CCNO", "FOXJ1", "PIFO", "RSPH1", "DKK1"),
    rownames(dat))
plotDots(dat, features = markers,
         group = "leiden_res0.125_annot", 
         swap_rownames = "gene_name", zlim = c(0, 7.5),
         colour = col(seq(0, 7.5, length.out = 64))) + 
    scale_size(limits=c(0.2, 1)) + 
    labs(x = "") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Marker genes from HECA paper (https://www.nature.com/articles/s41588-024-01873-w, Fig 2a).

```{r markers2}
#| fig.width: 8
#| fig.height: 12
#| warning: false

markers <- intersect(
    c("PAX8", "EPCAM", "FOXA2", "MMP7", "SOX9", "AXIN2", "TRH", "ALDH1A1",
      "CDH2", "KLK11", "SLC7A11", "DKK1", "PHLDA1", "KMO", "IHH", "EMID1", 
      "HPRT1", "SUFU", "OPRK1", "CBR3", "S100P", "SCGB2A2", "ABCG1", "PAEP", 
      "DPP4", "GPX3", "PTPRR", "FGF7", "FXYD2", "IL32", "WNT7A", "LGR5", 
      "MMP7", "TNF", "VTCN1", "CLDN22", "SULT1E1", "LEFTY1", "PTGS1", "IL6",
      "CCNO", "CDC20B", "MUC12", "PIFO", "FOXJ1", "TP73", "BPIFB1", "TFF3", 
      "MUC5B", "SAA1", "TP63", "SNCG", "KRT5", "PGR", "AR", "ESR1"), 
    rownames(dat))
plotDots(dat, features = markers,
         group = "leiden_res0.125_annot", 
         swap_rownames = "gene_name", zlim = c(0, 7.5),
         colour = col(seq(0, 7.5, length.out = 64))) + 
    scale_size(limits=c(0.2, 1)) + 
    labs(x = "") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

# Plot dimension reductions

```{r plot-tsne}
#| fig.width: 8
#| fig.height: 6

timepoint_cols <- c(
    "Horm Withdr 48h" = "grey40",
    "Post-breakdown 12h" = "#f3a9c0",
    "Post-breakdown 24h" = "#e27897",
    "Post-breakdown 48h" = "#cb3b63",
    "E2-diff 24h" = "#debaf5",
    "E2-diff 48h" = "#BE9FD2"
)
donor_cols <- c(
    "B050" = "#FF9966", 
    "B066" = "slateblue", 
    "B080" = "darkgreen"
)
seurat_cols <- c( 
    "S" = "gold",
    "G1" = "grey80",
    "G2M" = "brown"
) 

for (vrb in c("scDblFinder.class", "leiden_res0.125")) {
    print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                         colour_by = vrb, point_size = 0.5) + 
              guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
              theme_void())
}

print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                     colour_by = "leiden_res0.125_annot", point_size = 0.5) + 
          scale_color_manual(values = leiden_res0.125_annot_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())
print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                     colour_by = "timepoint", point_size = 0.5) + 
          scale_color_manual(values = timepoint_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())
print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                     colour_by = "donorCol", point_size = 0.5) + 
          scale_color_manual(values = donor_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())
print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                     colour_by = "SeuratCC", point_size = 0.5) + 
          scale_color_manual(values = seurat_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
          theme_void())


for (vrb in c("subsets_MT_percent", "detected", "intronic_fraction",
              "CCNO", "FOXJ1", "MKI67")) {
    print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                         colour_by = vrb, point_size = 0.5) + 
              theme_void())
}
```

# Transfer labels from tissue atlas

```{r transfer-labels}
#| fig.width: 8
#| fig.height: 8

set.seed(123)
atlas <- readRDS(tissue_atlas_sce)
ref_summed <- aggregateAcrossCells(atlas, id = atlas$leiden_res0.4_annot,
                                   use.assay.type = "logcounts",
                                   statistics = "mean")
pred <- SingleR(test = dat, ref = ref_summed, 
                labels = ref_summed$leiden_res0.4_annot)
plotScoreHeatmap(pred)
stopifnot(all(rownames(pred) == colnames(dat)))
dat$pred_from_invivo <- pred$labels
dat$pred_from_invivo_pruned <- pred$pruned.labels
dat$pred_from_invivo_delta <- pred$delta.next
```

# Visualize predicted labels

```{r plot-dimred}
print(plotReducedDim(dat, dimred = "UMAP_sketch", 
                     colour_by = "pred_from_invivo", point_size = 0.5) +
          scale_color_manual(values = atlas_leiden_res0.4_annot_cols, name = "") + 
          guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))))
```

# Plot cluster abundance over time

```{r}
#| warning: false
#| fig.height: 8
#| fig.width: 9.5

for (col in c("leiden_res0.125_annot", "pred_from_invivo")) {
    df <- data.frame(colData(dat)[, c("timepoint", col)]) |>
        # mutate(timepoint = factor(timepoint, 
        #                           levels = c("d8", "d9_12h", "d9_24h", 
        #                                      "d10", "d11", "d12"))) |>
        group_by(timepoint, .data[[col]]) |>
        tally() |>
        group_by(timepoint) |>
        mutate(perc = n / sum(n) * 100) |>
        group_by(.data[[col]]) |>
        mutate(scaledperc = perc / sum(perc)) |>
        ungroup()
    
    # by donor
    dfdonor <- data.frame(colData(dat)[, c("timepoint", "donor", col)]) |>
        group_by(timepoint, donor, .data[[col]]) |>
        tally() |>
        group_by(timepoint, donor) |>
        mutate(perc = n / sum(n) * 100) |>
        group_by(.data[[col]]) |>
        mutate(scaledperc = perc / sum(perc)) |>
        ungroup()
    
    if (col == "leiden_res0.125_annot") {
        df$cluster_group <- ifelse(grepl("ciliated", df[[col]]), 
                                   "ciliated", "non-ciliated")
    }
    gg <- ggplot(df, aes(x = timepoint, y = perc, color = .data[[col]])) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        theme_cowplot() + 
        labs(x = "Timepoint", y = "Relative abundance (%)",
             title = col) + 
        theme(legend.position = "bottom")
    if (col == "leiden_res0.125_annot") {
        gg <- gg + facet_wrap(~ cluster_group, ncol = 1) + 
            scale_color_manual(values = leiden_res0.125_annot_cols, name = "")
    } else {
        gg <- gg + 
            scale_color_manual(values = atlas_leiden_res0.4_annot_cols, name = "")
    }
    print(gg)
    
    ## One panel per cluster
    gg <- ggplot(df, aes(x = timepoint, y = perc, color = .data[[col]])) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        theme_cowplot() + 
        labs(x = "Timepoint", y = "Relative abundance (%)",
             title = col) + 
        facet_wrap(~ .data[[col]], scales = "free_y") + 
        theme(legend.position = "none",
              axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
    if (col == "leiden_res0.125_annot") {
        gg <- gg + scale_color_manual(values = leiden_res0.125_annot_cols, name = "")
    } else {
        gg <- gg + 
            scale_color_manual(values = atlas_leiden_res0.4_annot_cols, name = "")
    }
    print(gg)
    
    ## One panel per cluster, add lines for donors
    gg <- ggplot(bind_rows(df |> mutate(donor = "overall"), dfdonor), 
                 aes(x = timepoint, y = perc, group = donor, alpha = donor == "overall", 
                     linewidth = donor, color = .data[[col]])) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
        scale_linewidth_manual(values = c(`TRUE` = 2, `FALSE` = 0.75)) + 
        theme_cowplot() + 
        labs(x = "Timepoint", y = "Relative abundance (%)",
             title = col) + 
        facet_wrap(~ .data[[col]], scales = "free_y") + 
        theme(legend.position = "none",
              axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
    if (col == "leiden_res0.125_annot") {
        gg <- gg + scale_color_manual(values = leiden_res0.125_annot_cols, name = "")
    } else {
        gg <- gg + 
            scale_color_manual(values = atlas_leiden_res0.4_annot_cols, name = "")
    }
    print(gg)
    
    ## Bar plot
    gg <- ggplot(df, aes(x = timepoint, y = perc, fill = .data[[col]])) + 
        geom_col() + 
        theme_cowplot() + 
        labs(x = "Timepoint", y = "Relative abundance (%)", title = col) +
        coord_flip() + 
        theme(legend.position = "bottom")
    if (col == "leiden_res0.125_annot") {
        gg <- gg + scale_fill_manual(values = leiden_res0.125_annot_cols, name = "")
    } else {
        gg <- gg + 
            scale_fill_manual(values = atlas_leiden_res0.4_annot_cols, name = "")
    }
    print(gg)
}
```

# Find marker genes

First, we compare all clusters in a pairwise manner. The comparison is only 
performed for genes with a non-zero count in at least 
0.05% of the cells (corresponding to `r ceiling(0.05/100*ncol(dat))` of the 
`r ncol(dat)` cells in the data set). These gene IDs are also written to a 
file, for potential use as the 'universe' in over-representation analyses. 

```{r}
#| warning: false

frac_detected <- rowMeans(assay(dat, "counts") > 0)
genes_to_test <- rownames(dat)[frac_detected >= (0.05 / 100)]
write.table(genes_to_test, file = "organoid_timecourse_gene_universe.txt", 
            row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")

clc <- "leiden_res0.125_annot"
ptts <- pairwiseTTests(dat, groups = dat[[clc]], 
                       direction = "up", 
                       subset.row = genes_to_test)
# markers wrt all other clusters
markers <- combineMarkers(ptts$statistics, ptts$pairs, pval.type = "all")
# markers within lineage
keep_ciliated <- which(grepl("ciliated", ptts$pairs$first) & 
                           grepl("ciliated", ptts$pairs$second))
markers_ciliated <- combineMarkers(ptts$statistics[keep_ciliated], 
                                   ptts$pairs[keep_ciliated, ], 
                                   pval.type = "all")
keep_nonciliated <- which(!grepl("ciliated", ptts$pairs$first) & 
                              !grepl("ciliated", ptts$pairs$second))
markers_nonciliated <- combineMarkers(ptts$statistics[keep_nonciliated], 
                                      ptts$pairs[keep_nonciliated, ], 
                                      pval.type = "all")

## Write results to files
ttestdir <- paste0("pairwise_ttests_", clc)
if (!file.exists(ttestdir)) {
    dir.create(ttestdir, recursive = TRUE)
}
for (i in seq_len(nrow(ptts$pairs))) {
    df <- cbind(gene = rownames(ptts$statistics[[i]]), 
                ptts$statistics[[i]])
    df <- df[order(df$p.value, -df$logFC), ]
    filename <- paste0(clc, "_cluster_", ptts$pairs$first[i],
                       "_vs_", ptts$pairs$second[i], ".txt")
    write.table(df, file = file.path(ttestdir, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}

markerdir <- paste0("marker_genes_all_", clc)
if (!file.exists(markerdir)) {
    dir.create(markerdir, recursive = TRUE)
}
for (i in seq_along(markers)) {
    df <- cbind(gene = rownames(markers[[i]]), markers[[i]])
    filename <- paste0(clc, "_cluster_", names(markers)[i], "_markers.txt")
    write.table(df, file = file.path(markerdir, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}

markerdirlin <- paste0("marker_genes_within_lineage_all_", clc)
if (!file.exists(markerdirlin)) {
    dir.create(markerdirlin, recursive = TRUE)
}
for (i in seq_along(markers_ciliated)) {
    df <- cbind(gene = rownames(markers_ciliated[[i]]), 
                markers_ciliated[[i]])
    filename <- paste0(clc, "_cluster_", names(markers_ciliated)[i],
                       "_markers_within_lineage.txt")
    write.table(df, file = file.path(markerdirlin, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}
for (i in seq_along(markers_nonciliated)) {
    df <- cbind(gene = rownames(markers_nonciliated[[i]]), 
                markers_nonciliated[[i]])
    filename <- paste0(clc, "_cluster_", names(markers_nonciliated)[i],
                       "_markers_within_lineage.txt")
    write.table(df, file = file.path(markerdirlin, filename), quote = FALSE,
                sep = "\t", row.names = FALSE, col.names = TRUE)
}
```

# Save object

```{r}
saveRDS(dat, file = "organoid_scrnaseq_timecourse.rds")
```

# Session info

```{r}
sessionInfo()
```

