---
title: "Generating plots for paper (scRNA-seq time course)"
date: "`r Sys.Date()`"
format: 
    html:
        toc: true
        theme: united
embed-resources: true
---

# Load packages

```{r}
#| label: load-pkg

suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(scuttle)
    library(scater)
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(cowplot)
    library(ComplexHeatmap)
    library(circlize)
    library(ggtext)
    library(scales)
    library(ggh4x)
    library(swissknife)
    library(BiocParallel)
    library(Matrix)
    library(gprofiler2)
    library(forcats)
})

source("../params.R")
source("../helpers.R")

if (!file.exists("paper_figures")) {
    dir.create("paper_figures", showWarnings = FALSE, recursive = FALSE)
}
if (!file.exists("gprofiler")) {
    dir.create("gprofiler", showWarnings = FALSE, recursive = FALSE)
}
```

# Load data

```{r}
#| label: load-data

organ <- readRDS("organoid_scrnaseq_timecourse.rds")

# check for consistency
stopifnot(all(organ$leiden_res0.125_annot %in% sc_leiden_res0.125_annot_order))
stopifnot(all(organ$pred_from_invivo %in% tissue_leiden_res0.4_annot_order))

organ$leiden_res0.125_annot <- factor(as.character(organ$leiden_res0.125_annot),
                                      levels = sc_leiden_res0.125_annot_order)
organ$pred_from_invivo <- factor(as.character(organ$pred_from_invivo), 
                                 levels = tissue_leiden_res0.4_annot_order)

stopifnot(all(organ$timepoint %in% sc_timepoint_order))
organ$timepoint <- factor(organ$timepoint, 
                          levels = sc_timepoint_order)
```

# Numeric summaries

```{r}
# number of genes and cells
dim(organ)
```

# Figure 4B (UMAPs for time points)

```{r}
#| label: plot-umap-by-timepoint
#| fig.width: 22
#| fig.height: 4

x <- makePerCellDF(organ, use.coldata = TRUE, use.dimred = TRUE)

(umap_facet <- ggplot(x, aes(x = UMAP_sketch.1, y = UMAP_sketch.2)) + 
        geom_point(data = transform(x, timepoint = NULL), 
                   color = "grey95", alpha = 1, size = 0.5) + 
        geom_point(aes(color = timepoint), alpha = 0.5, size = 0.5) + 
        facet_wrap2(~ timepoint, nrow = 1, 
                    strip = strip_themed(clip = "off", background_x = elem_list_rect(
                        fill = alpha(sc_timepoint_colors[levels(x$timepoint)], 0.5)))) + 
        scale_color_manual(values = sc_timepoint_colors) + 
        theme_void() + 
        theme(axis.text = element_blank(),
              axis.title = element_blank(),
              strip.text = element_text(size = 20, hjust = 0.5, 
                                        margin = margin(0.15, 0, 0.15, 0, "cm")),
              legend.position = "none",
              panel.background = element_rect(fill = "white", color = NA),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line = element_line(color = "black"),
              strip.background = element_rect(fill = "white", color = "white")
        ))

png("paper_figures/UMAP_timepoints_separate.png", 
    height = 4, width = 22, units = "in", res = 500)
umap_facet
dev.off()
```

# Figure 4C (UMAP with clusters)

```{r}
#| label: plot-umap-clusters
#| fig.width: 7
#| fig.height: 4

(umap_clusters <- plotReducedDim(
    organ, dimred = "UMAP_sketch", 
    colour_by = "leiden_res0.125_annot", point_size = 0.5) + 
        scale_color_manual(values = sc_leiden_res0.125_annot_colors, name = "",
                           labels = sc_leiden_res0.125_annot_labels) + 
        guides(color = guide_legend(override.aes = list(size = 5, alpha = 1, 
                                                        shape = 15), 
                                    ncol = 2)) + 
        theme_void() + 
        theme(legend.text = element_markdown(size = 12)))
png("paper_figures/UMAP_organoids_clusters.png", 
    height = 4, width = 4, units = "in", res = 500)
umap_clusters + theme(legend.position = "none")
dev.off()
pdf("paper_figures/UMAP_organoids_clusters_legend.pdf", 
    height = 3, width = 5)
grid.newpage()
grid.draw(get_legend2(umap_clusters + 
                          theme(legend.position = "bottom",
                                legend.text = element_markdown(size = 12))))
dev.off()
pdf("paper_figures/UMAP_organoids_clusters.pdf", height = 4, width = 4)
umap_clusters + theme(legend.position = "none")
dev.off()
```

# Figure 4H (predictions from atlas)

```{r}
#| label: plot-umap-preds-invivo
#| fig.width: 7
#| fig.height: 4

(umap_preds <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "pred_from_invivo", point_size = 0.5) + 
        scale_color_manual(values = tissue_leiden_res0.4_annot_colors, name = "",
                           labels = tissue_leiden_res0.4_annot_labels) +  
        theme_void() + 
        theme(legend.text = element_markdown(size = 12),       
              legend.title = element_text(size = 12),      
              legend.key.size = unit(1.5, "lines")) + 
        guides(color = guide_legend(override.aes = list(size = 4), title = NULL))
)
png("paper_figures/UMAP_organoids_pred_from_invivo.png", 
    height = 4, width = 4, units = "in", res = 500)
umap_preds + theme(legend.position = "none")
dev.off()
pdf("paper_figures/UMAP_organoids_pred_from_invivo_legend.pdf", 
    height = 1.75, width = 9)
grid.newpage()
grid.draw(get_legend2(umap_preds + theme(legend.position = "bottom",
                                         legend.text = element_markdown(size = 12))))
dev.off()
```

# Figure 4E (dot plot)

```{r}
#| label: plot-dotplot-timecourse-markers
#| fig.width: 20
#| fig.height: 6

markers <- intersect(sc_timecourse_markers, rownames(organ))

(dotplot_norm <- .makeDotPlot(
    dat = organ, markers = markers, 
    cluster_column = "leiden_res0.125_annot", 
    cluster_order = sc_leiden_res0.125_annot_order, 
    cluster_colors = sc_leiden_res0.125_annot_colors, 
    cluster_labels = sc_leiden_res0.125_annot_labels,
    transpose = TRUE, 
    split_by_ciliated = TRUE))

pdf("paper_figures/dotplot_organoids_normalized.pdf", height = 6, width = 20)
dotplot_norm
dev.off()
png("paper_figures/dotplot_organoids_normalized.png", height = 6, width = 20, 
    unit = "in", res = 500)
dotplot_norm
dev.off()
```

# Dot plot of WNT ligands and FZD receptors

```{r}
#| label: plot-dotplot-timecourse-wnt-fzd
#| fig.width: 7
#| fig.height: 6

fzd <- paste0("FZD", 1:10)

(dotplot_norm_fzd <- .makeDotPlot(
    dat = organ, markers = fzd, 
    cluster_column = "leiden_res0.125_annot", 
    cluster_order = sc_leiden_res0.125_annot_order, 
    cluster_colors = sc_leiden_res0.125_annot_colors, 
    cluster_labels = sc_leiden_res0.125_annot_labels, 
    transpose = TRUE, 
    split_by_ciliated = FALSE))

pdf("paper_figures/dotplot_organoids_fzd_normalized.pdf", height = 6, width = 7)
dotplot_norm_fzd
dev.off()
png("paper_figures/dotplot_organoids_fzd_normalized.png", height = 6, width = 7, 
    unit = "in", res = 500)
dotplot_norm_fzd
dev.off()
```

# Figures 4D + 4G + 4I, Extended Data 9G (relative abundance over time)

## Time course clusters

```{r}
#| label: plot-rel-abundance-clusters
#| fig.width: 9
#| fig.height: 6

df <- data.frame(colData(organ)[, c("timepoint", "leiden_res0.125_annot")]) |>
    group_by(timepoint, leiden_res0.125_annot) |>
    tally() |>
    group_by(timepoint) |>
    mutate(perc = n / sum(n) * 100) |>
    ungroup() |>
    complete(timepoint, leiden_res0.125_annot, fill = list(n = 0, perc = 0)) |>
    mutate(cluster_group = ifelse(grepl("ciliated", leiden_res0.125_annot), 
                                  "ciliated", "non-ciliated"))

dfdonor <- data.frame(colData(organ)[, c("timepoint", "donor", "leiden_res0.125_annot")]) |>
    group_by(timepoint, donor, leiden_res0.125_annot) |>
    tally() |>
    group_by(timepoint, donor) |>
    mutate(perc = n / sum(n) * 100) |>
    ungroup() |>
    complete(timepoint, donor, leiden_res0.125_annot, fill = list(n = 0, perc = 0))

# timepoint vs. relative abundance
# ... ciliated
(plot_relab_cil <- ggplot(df |> dplyr::filter(cluster_group == "ciliated") |>
                              droplevels(), 
                          aes(x = timepoint, y = perc, color = leiden_res0.125_annot)) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        theme_cowplot() + 
        labs(x = "", y = "Relative\nabundance (%)",
             title = "") + 
        theme(legend.position = "bottom",
              legend.text = element_markdown(),
              axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16),
              strip.text = element_text(size = 16, margin = margin(0.2, 0, 0.2, 0, "cm")),
              strip.background = element_rect(fill = "white", color = "white")) + 
        facet_wrap(~ cluster_group, ncol = 1) + 
        scale_color_manual(values = sc_leiden_res0.125_annot_colors, name = "",
                           labels = sc_leiden_res0.125_annot_labels))

pdf(paste0("paper_figures/relative_abundance_leiden_res0.125_annot_ciliated.pdf"), 
    height = 4.5, width = 6)
print(plot_relab_cil + theme(legend.position = "none"))
dev.off()
pdf("paper_figures/relative_abundance_leiden_res0.125_annot_ciliated_legend.pdf", 
    height = 3, width = 4)
grid.newpage()
grid.draw(get_legend2(plot_relab_cil + 
                          theme(legend.position = "bottom",
                                legend.text = element_markdown(size = 12)) + 
                          guides(color = guide_legend(ncol = 1))))
dev.off()

# ... non-ciliated
(plot_relab_noncil <- ggplot(df |> dplyr::filter(cluster_group == "non-ciliated") |>
                                 droplevels(), 
                             aes(x = timepoint, y = perc, color = leiden_res0.125_annot)) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        theme_cowplot() + 
        labs(x = "", y = "Relative\nabundance (%)",
             title = "") + 
        theme(legend.position = "bottom",
              legend.text = element_markdown(),
              axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16),
              strip.text = element_text(size = 16, margin = margin(0.2, 0, 0.2, 0, "cm")),
              strip.background = element_rect(fill = "white", color = "white")) + 
        facet_wrap(~ cluster_group, ncol = 1) + 
        scale_color_manual(values = sc_leiden_res0.125_annot_colors, name = "",
                           labels = sc_leiden_res0.125_annot_labels))

pdf(paste0("paper_figures/relative_abundance_leiden_res0.125_annot_nonciliated.pdf"), 
    height = 4.5, width = 6)
print(plot_relab_noncil + theme(legend.position = "none"))
dev.off()
pdf("paper_figures/relative_abundance_leiden_res0.125_annot_nonciliated_legend.pdf", 
    height = 2, width = 4.5)
grid.newpage()
grid.draw(get_legend2(plot_relab_noncil + 
                          theme(legend.position = "bottom",
                                legend.text = element_markdown(size = 12)) + 
                          guides(color = guide_legend(ncol = 2))))
dev.off()
```

```{r}
#| label: plot-rel-abundance-clusters-facet
#| fig.width: 14
#| fig.height: 10

# one panel per cluster
(plot_relab_facet <- ggplot(
    bind_rows(df |> mutate(donor = "overall"), dfdonor), 
    aes(x = timepoint, y = perc, group = donor, 
        alpha = donor == "overall", 
        linewidth = donor == "overall", color = leiden_res0.125_annot)) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint))) + 
        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
        scale_linewidth_manual(values = c(`TRUE` = 2, `FALSE` = 1)) + 
        theme_cowplot() + 
        labs(x = "", y = "Relative abundance (%)",
             title = "") + 
        scale_color_manual(values = sc_leiden_res0.125_annot_colors, name = "",
                           labels = sc_leiden_res0.125_annot_labels) + 
        facet_wrap(~ leiden_res0.125_annot, scales = "free_y",
                   labeller = labeller(leiden_res0.125_annot = sc_leiden_res0.125_annot_labels)) + 
        guides(alpha = "none", linewidth = "none") + 
        theme(legend.position = "none",
              legend.text = element_markdown(),
              axis.text.x = element_text(angle = 45, hjust = 1),
              strip.text = element_markdown(size = 16),
              strip.background = element_rect(fill = "white", color = "white")))

pdf(paste0("paper_figures/relative_abundance_cluster_leiden_res0.125_annot.pdf"), 
    height = 9, width = 14)
print(plot_relab_facet)
dev.off()
```

```{r}
#| label: plot-rel-abundance-clusters-barplot
#| fig.width: 8
#| fig.height: 6

# bar plot
(plotbar <- ggplot(df, aes(x = fct_rev(timepoint), y = perc, fill = leiden_res0.125_annot)) + 
        geom_col() + 
        theme_cowplot() + 
        labs(x = "", y = "Relative abundance (%)", title = "") +
        coord_flip() + 
        theme(legend.position = "bottom",
              legend.text = element_markdown(),
              axis.text = element_text(size = 20),
              axis.title = element_text(size = 20)) + 
        scale_fill_manual(values = sc_leiden_res0.125_annot_colors, name = "",
                          labels = sc_leiden_res0.125_annot_labels))
pdf("paper_figures/relative_abundance_barplot_leiden_res0.125_annot.pdf", 
    height = 5, width = 8)
plotbar + theme(legend.position = "none")
dev.off()
png("paper_figures/relative_abundance_barplot_leiden_res0.125_annot.png", 
    height = 5, width = 8, unit = "in", res = 400)
plotbar + theme(legend.position = "none")
dev.off()
pdf("paper_figures/relative_abundance_barplot_leiden_res0.125_annot_legend.pdf", 
    height = 1.75, width = 9)
grid.newpage()
grid.draw(get_legend2(plotbar + theme(legend.position = "bottom",
                                      legend.text = element_markdown(size = 12))))
dev.off()
```

## Predicted labels from atlas

```{r}
#| label: plot-rel-abundance-pred
#| fig.width: 9
#| fig.height: 6

df <- data.frame(colData(organ)[, c("timepoint", "pred_from_invivo")]) |>
    group_by(timepoint, pred_from_invivo) |>
    tally() |>
    group_by(timepoint) |>
    mutate(perc = n / sum(n) * 100) |>
    ungroup() |>
    complete(timepoint, pred_from_invivo, fill = list(n = 0, perc = 0))

dfdonor <- data.frame(colData(organ)[, c("timepoint", "donor", "pred_from_invivo")]) |>
    group_by(timepoint, donor, pred_from_invivo) |>
    tally() |>
    group_by(timepoint, donor) |>
    mutate(perc = n / sum(n) * 100) |>
    ungroup() |>
    complete(timepoint, donor, pred_from_invivo, fill = list(n = 0, perc = 0))

# timepoint vs. relative abundance
(plot_relab <- ggplot(df, aes(x = timepoint, y = perc, color = pred_from_invivo)) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint)), linewidth = 1) + 
        theme_cowplot() + 
        labs(x = "", y = "Relative\nabundance (%)",
             title = "") + 
        theme(legend.position = "bottom",
              legend.text = element_markdown(),
              axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16),
              strip.background = element_rect(fill = "white", color = "black")) + 
        scale_color_manual(values = tissue_leiden_res0.4_annot_colors, name = "",
                           labels = tissue_leiden_res0.4_annot_labels))

pdf(paste0("paper_figures/relative_abundance_pred_from_invivo.pdf"), 
    height = 4, width = 6)
print(plot_relab + theme(legend.position = "none"))
dev.off()
pdf("paper_figures/relative_abundance_pred_from_invivo_legend.pdf", 
    height = 1.75, width = 11)
grid.newpage()
grid.draw(get_legend2(plot_relab + theme(legend.position = "bottom",
                                         legend.text = element_markdown(size = 12))))
dev.off()
```

```{r}
#| label: plot-rel-abundance-pred-facet
#| fig.width: 14
#| fig.height: 9

# one panel per cluster
(plot_relab_facet <- ggplot(
    bind_rows(df |> mutate(donor = "overall"), dfdonor), 
    aes(x = timepoint, y = perc, group = donor, 
        alpha = donor == "overall", 
        linewidth = donor == "overall", color = pred_from_invivo)) + 
        geom_point() + 
        geom_line(aes(x = as.numeric(timepoint))) + 
        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
        scale_linewidth_manual(values = c(`TRUE` = 2, `FALSE` = 1)) + 
        theme_cowplot() + 
        labs(x = "", y = "Relative abundance (%)",
             title = "") + 
        theme(legend.position = "none",
              axis.text.x = element_text(angle = 45, hjust = 1)) + 
        scale_color_manual(values = tissue_leiden_res0.4_annot_colors, name = "",
                           labels = tissue_leiden_res0.4_annot_labels) + 
        facet_wrap(~ pred_from_invivo, scales = "free_y", 
                   labeller = labeller(pred_from_invivo = tissue_leiden_res0.4_annot_labels)) + 
        guides(alpha = "none", linewidth = "none") + 
        theme(strip.text.x = element_markdown(size = 14),
              strip.text.y = element_markdown(size = 14),
              strip.background = element_rect(fill = "white", color = "white")))

pdf(paste0("paper_figures/relative_abundance_cluster_pred_from_invivo.pdf"), 
    height = 9, width = 14)
print(plot_relab_facet)
dev.off()
```

```{r}
#| label: plot-rel-abundance-pred-barplot
#| fig.width: 8
#| fig.height: 6

# bar plot
(plotbar <- ggplot(df, aes(x = timepoint, y = perc, fill = pred_from_invivo)) + 
        geom_col() + 
        theme_cowplot() + 
        labs(x = "", y = "Relative abundance (%)", title = "") +
        coord_flip() + 
        theme(legend.position = "bottom",
              legend.text = element_markdown(),
              axis.text = element_text(size = 20),
              axis.title = element_text(size = 20)) + 
        scale_fill_manual(values = tissue_leiden_res0.4_annot_colors, name = "",
                          labels = tissue_leiden_res0.4_annot_labels))
pdf("paper_figures/relative_abundance_barplot_pred_from_invivo.pdf", 
    height = 5, width = 8)
plotbar + theme(legend.position = "none")
dev.off()
pdf("paper_figures/relative_abundance_barplot_pred_from_invivo_legend.pdf", 
    height = 1.75, width = 11)
grid.newpage()
grid.draw(get_legend2(plotbar + theme(legend.position = "bottom",
                                      legend.text = element_markdown(size = 12))))
dev.off()
```


# Extended Data 9 (UMAPs)

```{r}
#| label: plot-umap-qc
#| fig.width: 6
#| fig.height: 4

(umap_donor <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "donorCol",
    point_size = 0.5) +
        scale_color_manual(values = donor_colors) + 
        guides(color = guide_legend(override.aes = list(size = 4),
                                    title = "Donor")) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines")))
png("paper_figures/UMAP_organoids_donorCol.png", 
    height = 4, width = 7, units = "in", res = 500)
plot_grid(
    umap_donor + theme(legend.position = "none"),
    get_legend2(umap_donor),
    nrow = 1, rel_widths = c(4, 3)
)
dev.off()

(umap_cc <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "SeuratCC",
    point_size = 0.5) +
        scale_color_manual(values = cc_colors) + 
        guides(color = guide_legend(override.aes = list(size = 4),
                                    title = "Cell cycle")) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines")))
png("paper_figures/UMAP_organoids_seuratCC.png", 
    height = 4, width = 7, units = "in", res = 500)
plot_grid(
    umap_cc + theme(legend.position = "none"),
    get_legend2(umap_cc),
    nrow = 1, rel_widths = c(4, 3)
)
dev.off()

(umap_tricycle <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "tricyclePosition",
    point_size = 0.5) +   
        guides(color = guide_colorbar(title = "tricycle\nposition")) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines")))
png("paper_figures/UMAP_organoids_tricyclePosition.png", 
    height = 4, width = 7, units = "in", res = 500)
plot_grid(
    umap_tricycle + theme(legend.position = "none"),
    get_legend2(umap_tricycle),
    nrow = 1, rel_widths = c(4, 3)
)
dev.off()

(umap_intronic <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "intronic_fraction",
    point_size = 0.5) +   
        guides(color = guide_colorbar(title = "intronic\nfraction")) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines")))
png("paper_figures/UMAP_organoids_intronic_fraction.png", 
    height = 4, width = 7, units = "in", res = 500)
plot_grid(
    umap_intronic + theme(legend.position = "none"),
    get_legend2(umap_intronic),
    nrow = 1, rel_widths = c(4, 3)
)
dev.off()

(umap_mt <- plotReducedDim(
    organ, "UMAP_sketch", colour_by = "subsets_MT_percent",
    point_size = 0.5) +   
        guides(color = guide_colorbar(title = "% MT")) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines")))
png("paper_figures/UMAP_organoids_subsets_MT_percent.png", 
    height = 4, width = 7, units = "in", res = 500)
plot_grid(
    umap_mt + theme(legend.position = "none"),
    get_legend2(umap_mt),
    nrow = 1, rel_widths = c(4, 3)
)
dev.off()
```

# Figure 4F (UMAPs)

```{r}
#| label: plot-umap-timecourse-genes
#| fig.width: 6
#| fig.height: 4

for (gn in c("WNT7A", "E2F7")) {
    umap_gn <- plotReducedDim(
        organ, "UMAP_sketch", colour_by = gn,
        point_size = 1) + 
        scale_color_gradientn(colors = c("#d9d9d9", "#f4a582", "#d6604d","#8b0000", '#5c0000')) +
        labs(color = gn) + 
        theme_void() + 
        theme(axis.title = element_blank(),
              axis.text = element_blank(),
              legend.text = element_text(size = 20),      
              legend.title = element_text(size = 20),      
              legend.key.size = unit(1.5, "lines"))
    print(umap_gn)
    
    png(paste0("paper_figures/UMAP_organoids_", gn, ".png"), 
        height = 4, width = 6, units = "in", res = 500)
    print(plot_grid(
        umap_gn + theme(legend.position = "none"),
        get_legend2(umap_gn),
        nrow = 1, rel_widths = c(4, 2)
    ))
    dev.off()
}
```

# Luminal early signature over time

Use the top 100 marker genes for the luminal early cluster to calculate a 
gene set score for each cell. 

```{r}
#| label: fig-lum-early-signature
#| fig.width: 7
#| fig.height: 5

luminal_early_markers <- read.delim("../scRNAseq_invivo_endometrium_epithelial_cell_atlas/marker_genes_all_leiden_res0.4_annot/leiden_res0.4_annot_cluster_luminal early_markers.txt") |>
    arrange(p.value) |>
    slice_head(n = 100) |>
    pull(gene)

set.seed(478L)
organ$lum_early_score <- swissknife::normGenesetExpression(
    organ, genes = luminal_early_markers, 
    BPPARAM = MulticoreParam(workers = 8L, RNGseed = 478L))
(ggrep <- ggplot(as.data.frame(colData(organ)), 
                 aes(x = timepoint, y = lum_early_score,
                     fill = timepoint)) + 
        geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
        labs(x = "", y = "Regeneration associated\nluminal gene set score") + 
        scale_fill_manual(values = sc_timepoint_colors) + 
        theme_bw(base_size = 14) + 
        theme(legend.position = "none",
              axis.text.x = element_text(size = 16, angle = 45, hjust = 1, vjust = 1),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16)))

pdf("paper_figures/luminal_early_signature.pdf", width = 7, height = 5)
ggrep
dev.off()
```

Instead of the top 100 marker genes for the luminal early cluster, we here use 
the top 100 upregulated genes in the luminal early cluster when compared to the 
luminal late cluster. 

```{r}
#| label: fig-lum-early-vs-late-signature
#| fig.width: 7
#| fig.height: 5.5

luminal_early_vs_late_markers <- read.delim("../scRNAseq_invivo_endometrium_epithelial_cell_atlas/pairwise_ttests_leiden_res0.4_annot/leiden_res0.4_annot_cluster_luminal early_vs_luminal late.txt") |>
    arrange(p.value) |>
    slice_head(n = 100) |>
    pull(gene)

set.seed(478L)
organ$lum_early_vs_late_score <- swissknife::normGenesetExpression(
    organ, genes = luminal_early_vs_late_markers, 
    BPPARAM = MulticoreParam(workers = 8L, RNGseed = 478L))
(ggrep3 <- ggplot(as.data.frame(colData(organ)), 
                 aes(x = timepoint, y = lum_early_vs_late_score,
                     fill = timepoint)) + 
        geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
        labs(x = "", y = "Regeneration vs receptivity\nassociated luminal gene set score") + 
        scale_fill_manual(values = sc_timepoint_colors) + 
        theme_bw(base_size = 14) + 
        theme(legend.position = "none",
              axis.text.x = element_text(size = 16, angle = 45, hjust = 1, vjust = 1),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16)))

pdf("paper_figures/luminal_early_vs_late_signature.pdf", width = 7, height = 5.5)
ggrep3
dev.off()
```

# Luminal late signature over time

```{r}
#| label: fig-lum-late-signature
#| fig.width: 7
#| fig.height: 5

luminal_late_markers <- read.delim("../scRNAseq_invivo_endometrium_epithelial_cell_atlas/marker_genes_all_leiden_res0.4_annot/leiden_res0.4_annot_cluster_luminal late_markers.txt") |>
    arrange(p.value) |>
    slice_head(n = 100) |>
    pull(gene)

set.seed(478L)
organ$lum_late_score <- swissknife::normGenesetExpression(
    organ, genes = luminal_late_markers, 
    BPPARAM = MulticoreParam(workers = 8L, RNGseed = 478L))
(ggrep2 <- ggplot(as.data.frame(colData(organ)), 
                  aes(x = timepoint, y = lum_late_score,
                      fill = timepoint)) + 
        geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
        labs(x = "", y = "Receptivity associated\nluminal gene set score") + 
        scale_fill_manual(values = sc_timepoint_colors) + 
        theme_bw(base_size = 14) + 
        theme(legend.position = "none",
              axis.text.x = element_text(size = 16, angle = 45, hjust = 1, vjust = 1),
              axis.text.y = element_text(size = 16),
              axis.title = element_text(size = 16)))

pdf("paper_figures/luminal_late_signature.pdf", width = 7, height = 5)
ggrep2
dev.off()
```

# Overrepresentation analysis

```{r}
#| label: gprofiler
#| fig.height: 8
#| fig.width: 7

# tested genes
timecourse_gene_bg <- read.delim("organoid_timecourse_gene_universe.txt", 
                                 header = FALSE)[, 1]

clusters_to_consider <- c("stress responsive",
                          "wound responsive 12h",
                          "wound responsive 24h",
                          "immunomodulatory",
                          "OxPhos high", 
                          "cycling",
                          "hormone responsive")


gprofiler2::set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e112_eg59_p19")
for (cl in clusters_to_consider) {
    mgenes <- read.delim(paste0(
        "marker_genes_within_lineage_all_leiden_res0.125_annot/leiden_res0.125_annot_cluster_", 
        cl, "_markers_within_lineage.txt")) |>
        arrange(p.value) |>
        slice_head(n = 100) |>
        pull(gene)
    
    gostres <- gost(
        query = mgenes, 
        organism = "hsapiens",
        ordered_query = FALSE, 
        multi_query = FALSE, 
        significant = TRUE, 
        exclude_iea = TRUE, 
        measure_underrepresentation = FALSE, 
        evcodes = FALSE, 
        user_threshold = 0.05, 
        correction_method = "fdr",
        domain_scope = "custom",
        custom_bg = timecourse_gene_bg, 
        numeric_ns = "", 
        sources = c("GO:BP"),
        as_short_link = FALSE, 
        highlight = TRUE
    )
    print(head(gostres$result |>
                   dplyr::select(term_name, p_value, term_size, query_size, intersection_size),
               n = 15) |>
              mutate(p_value = signif(p_value, digits = 3)))
    write.table(gostres$result |>
                    dplyr::select(term_name, p_value, term_size, query_size, intersection_size),
                file = paste0("gprofiler/", sub(" ", "_", cl), "_markers_gobp.tsv"),
                row.names = FALSE, quote = FALSE, col.names = TRUE, sep = "\t")
}

plotdf <- do.call(bind_rows, lapply(clusters_to_consider, function(cl) {
    read.delim(paste0("gprofiler/", sub(" ", "_", cl), "_markers_gobp.tsv")) |>
        slice_head(n = 5) |>
        mutate(cluster = cl) |>
        mutate(term_size = as.numeric(term_size),
               p_value = as.numeric(p_value))
})) |>
    mutate(cluster2 = sub("immunomodulatory", "immuno-\nmodulatory", 
                          gsub(" ", "\n", cluster))) |>
    mutate(cluster = factor(cluster, levels = rev(clusters_to_consider))) |>
    mutate(cluster2 = factor(cluster2, levels = sub("immunomodulatory", "immuno-\nmodulatory",
                                                    gsub(" ", "\n", levels(cluster)))))
(ggprof <- ggplot(plotdf, aes(x = -log10(p_value), y = fct_reorder(term_name, -p_value))) + 
    geom_point(size = 4, aes(color = cluster)) + 
    scale_color_manual(values = sc_leiden_res0.125_annot_colors, 
                       labels = sc_leiden_res0.125_annot_labels) + 
    facet_grid2(cluster2 ~ .,
                strip = strip_themed(clip = "off", background_y = elem_list_rect(
                    fill = alpha(sc_leiden_res0.125_annot_colors[levels(plotdf$cluster)], 0.5))), 
                scales = "free_y") + 
    labs(x = "-log10 FDR", y = "") + 
    theme_cowplot() + 
    theme(legend.position = "none"))
pdf("paper_figures/gprofiler_selclusters.pdf", width = 7, height = 8)
ggprof
dev.off()
```

# Save count matrix and cell type annotations

```{r}
#| eval: false

if (!file.exists("upload_data")) {
    dir.create("upload_data", showWarnings = FALSE, recursive = FALSE)
}
writeMM(assay(organ, "counts"), 
        file = "upload_data/ivmc_scrnaseq_counts.mtx")
R.utils::gzip(filename = "upload_data/ivmc_scrnaseq_counts.mtx", 
              compression = 9, overwrite = TRUE, remove = TRUE)
write.table(rownames(organ), 
            file = "upload_data/ivmc_scrnaseq_genes.tsv", 
            row.names = FALSE, col.names = FALSE, sep = "\t", quote = FALSE)
R.utils::gzip(filename = "upload_data/ivmc_scrnaseq_genes.tsv", 
              compression = 9, overwrite = TRUE, remove = TRUE)
write.table(colnames(organ), 
            file = "upload_data/ivmc_scrnaseq_cells.tsv", 
            row.names = FALSE, col.names = FALSE, sep = "\t", quote = FALSE)
R.utils::gzip(filename = "upload_data/ivmc_scrnaseq_cells.tsv", 
              compression = 9, overwrite = TRUE, remove = TRUE)

write.table(
    cbind(colData(organ)[, c("cell", "leiden_res0.125_annot", 
                             "pred_from_invivo", "timepoint")], 
          reducedDim(organ, "UMAP_sketch")[, 1:2]),
    file = "upload_data/ivmc_scrnaseq_annot.tsv", 
    row.names = FALSE, col.names = TRUE, sep = "\t")
R.utils::gzip(filename = "upload_data/ivmc_scrnaseq_annot.tsv", 
              compression = 9, overwrite = TRUE, remove = TRUE)
```

# Session info

```{r}
#| label: session-info

# gProfiler/GO version
get_version_info(organism = "hsapiens")$gprofiler_version
get_version_info(organism = "hsapiens")$sources$`GO:MF`

sessionInfo()
```

